---
title: "PCA"
author: "JL"
date: "`r Sys.Date()`"
output: html_document
---
```{r}
library(here)
library(tidyverse)
library(patchwork)
library(ggrepel)
library(corrplot)
library(GGally)
```

```{r}
traits <- read.csv(here("Data", "cleaned", "BC_bird_traits.csv"))
traits$habitat_density[traits$habitat_density == "dense"] <- 1
traits$habitat_density[traits$habitat_density == "semiopen"] <- 2
traits$habitat_density[traits$habitat_density == "open"] <- 3
traits <- traits %>% 
    mutate(cavity_nester = ifelse(nest_category == "cavity", 1, 0),
   #        ground_nester = ifelse(most_common_nesting_location == "ground", 1, 0),
   #        shrub_nester = ifelse(most_common_nesting_location == "shrub", 1, 0)) %>%
    habitat_density = as.numeric(habitat_density)) 
# old_growth <-  c("American Three-toed Woodpecker", "Bay-breasted Warbler", "Black-capped Chickadee", "Boreal Chickadee", "Boreal Owl", "Brown Creeper", "Canada Warbler", "Chestnut-backed Chickadee", "Dusky Grouse", "Golden-crowned Kinglet", "Great Horned Owl", "Hairy Woodpecker", "Hammond's Flycatcher", "Hermit Thrush", "Lewis's Woodpecker", "American Goshawk", "Northern Waterthrush", "Western Flycatcher", "Pileated Woodpecker", "Red-breasted Nuthatch", "Red-breasted Sapsucker", "Ruby-crowned Kinglet", "Spruce Grouse", "Swainson's Thrush", "Townsend's Warbler", "Varied Thrush", "Vaux's Swift", "White-breasted Nuthatch")
# missing_names <- old_growth[!(old_growth %in% traits$common_name)] #do all these names appear in the BC list?
traits2 <- traits

##
positive <- c("American Crow", "American Goldfinch", "American Kestrel", "American Robin", "American Three-toed Woodpecker", "Baltimore Oriole", "Black-backed Woodpecker", "Black-billed Magpie", "Black-capped Chickadee", "Black-headed Grosbeak", "Black-throated Gray Warbler", "Brewer's Blackbird", "Brewer's Sparrow", "Brown Creeper", "Brown-headed Cowbird", "Canada Goose", "Canada Jay", "Cassin's Finch", "Cedar Waxwing", "Chestnut-backed Chickadee", "Chipping Sparrow", "Clay-colored Sparrow", "Common Yellowthroat", "Connecticut Warbler", "Dark-eyed Junco", "Downy Woodpecker", "Dusky Flycatcher", "Eastern Kingbird", "Evening Grosbeak", "Fox Sparrow", "Golden-crowned Kinglet", "Great Horned Owl", "Hairy Woodpecker", "Hammond's Flycatcher", "Hermit Thrush", "Lazuli Bunting", "Least Flycatcher", "Lewis's Woodpecker", "Lincoln's Sparrow", "MacGillivray's Warbler", "Mountain Bluebird", "Mountain Chickadee", "Nashville Warbler", "Northern Flicker", "Northern Hawk Owl", "Northern House Wren", "Northern Pygmy-Owl", "Olive-sided Flycatcher", "Pacific Wren", "Palm Warbler", "Pine Siskin", "Red-breasted Nuthatch", "Red-winged Blackbird", "Rock Wren", "Rufous Hummingbird", "Savannah Sparrow", "Say's Phoebe", "Song Sparrow", "Spotted Towhee", "Steller's Jay", "Swamp Sparrow", "Tennessee Warbler", "Townsend's Solitaire", "Tree Swallow", "Turkey Vulture", "Veery", "Warbling Vireo", "Western Bluebird", "Western Screech-Owl", "Western Tanager", "Western Wood-Pewee", "White-breasted Nuthatch", "White-crowned Sparrow", "White-throated Sparrow", "Winter Wren", "Yellow-bellied Sapsucker", "Yellow-rumped Warbler")
# "Broad-tailed Hummingbird","Eastern Bluebird", "Field Sparrow", "Prairie Warbler",  "White-headed Woodpecker","Red-headed Woodpecker",
missing_names <- positive[!(positive %in% traits$common_name)] #do all these names appear in the BC list?


negative <- c("Alder Flycatcher", "Barred Owl", "Bay-breasted Warbler", "Black-and-white Warbler", "Black-headed Grosbeak", "Black-throated Gray Warbler", "Boreal Chickadee", "Brown Creeper", "Canada Jay", "Cape May Warbler", "Cassin's Vireo", "Chestnut-backed Chickadee", "Common Goldeneye", "Common Raven", "Common Yellowthroat", "Dark-eyed Junco", "Dusky Flycatcher", "Golden-crowned Kinglet","Great Horned Owl",  "Hermit Thrush",  "Hutton's Vireo", "MacGillivray's Warbler", "Mountain Chickadee", "Nashville Warbler",  "Northern Saw-whet Owl", "Northern Waterthrush", "Orange-crowned Warbler", "Ovenbird", "Pacific Wren", "Pileated Woodpecker", "Pine Grosbeak", "Pine Siskin",  "Philadelphia Vireo", "Purple Finch", "Pygmy Nuthatch", "Red Crossbill", "Red-breasted Nuthatch", "Red-breasted Sapsucker", "Red-eyed Vireo", "Red-naped Sapsucker", "Ruffed Grouse", "Ruby-crowned Kinglet", "Sooty Grouse", "Steller's Jay", "Swainson's Thrush", "Swamp Sparrow", "Townsend's Warbler", "Vaux's Swift", "Warbling Vireo", "Western Tanager", "Western Wood-Pewee", "White-winged Crossbill", "Williamson's Sapsucker", "Wilson's Warbler", "Winter Wren",  "Yellow Warbler", "Yellow-rumped Warbler"
)
#"Broad-tailed Hummingbird", "Eastern Bluebird", "Field Sparrow","Prairie Warbler","Red-headed Woodpecker","White-headed Woodpecker",  "Blue-gray Gnatcatcher", "Great Crested Flycatcher", "Greater Sage-Grouse","Hooded Warbler","Northern Cardinal","Pinyon Jay","Wood Thrush",
missing_names <- negative[!(negative %in% traits$common_name)] #do all these names appear in the BC list?

null <- c("American Goldfinch", "American Kestrel", "American Robin", "American Wigeon", "Barrow's Goldeneye", "Black-capped Chickadee", "Black-headed Grosbeak", "Brewer's Blackbird", "Brown Creeper", "Brown-headed Cowbird", "Bufflehead", "Calliope Hummingbird", "Canvasback", "Cassin's Finch", "Cassin's Vireo", "Cedar Waxwing", "Chipping Sparrow", "Clark's Nutcracker", "Common Goldeneye", "Common Loon", "Common Merganser", "Common Nighthawk", "Common Raven", "Common Yellowthroat", "Dark-eyed Junco", "Downy Woodpecker", "Dusky Flycatcher", "Eastern Phoebe", "Evening Grosbeak", "Fox Sparrow",  "Green-winged Teal", "Hairy Woodpecker", "Hammond's Flycatcher", "Hermit Thrush", "House Finch", "Lesser Scaup", "Lincoln's Sparrow", "MacGillivray's Warbler", "Mallard", "Marsh Wren", "Mountain Chickadee", "Mourning Dove", "Nashville Warbler", "Northern Pintail", "Northern Shoveler", "Northern Waterthrush", "Olive-sided Flycatcher", "Orange-crowned Warbler", "Pacific Wren", "Pileated Woodpecker", "Pine Grosbeak", "Pine Siskin", "Pygmy Nuthatch", "Red Crossbill", "Red-breasted Merganser", "Red-breasted Nuthatch", "Red-breasted Sapsucker", "Red-naped Sapsucker", "Red-tailed Hawk", "Ring-necked Duck", "Rock Wren", "Ruby-crowned Kinglet", "Ruffed Grouse", "Rufous Hummingbird", "Song Sparrow",  "Steller's Jay", "Surf Scoter", "Swainson's Thrush", "Townsend's Solitaire", "Townsend's Warbler", "Varied Thrush", "Violet-green Swallow", "Warbling Vireo", "Western Tanager", "White-breasted Nuthatch", "White-crowned Sparrow", "White-winged Scoter", "Williamson's Sapsucker", "Willow Flycatcher", "Wilson's Warbler", "Winter Wren", "Yellow Warbler", "Yellow-bellied Flycatcher", "Yellow-rumped Warbler")
 #"Greater Scaup","Spotted Sandpiper",
missing_names <- null[!(null %in% traits$common_name)] #do all these names appear in the BC list?

traits$positive <- ifelse(traits$common_name %in% positive, 1, NA)
traits$negative <- ifelse(traits$common_name %in% negative, 1, NA)
traits$null <- ifelse(traits$common_name %in% null, 1, NA)


traits <- traits %>%
    filter(!if_all(c(positive, negative, null), is.na)) %>% 
    mutate(across(c(positive, negative, null), ~replace_na(.x, 0)))



```

```{r CORRELATION TEST}
trait.pca <- traits %>% 
  select(common_name, where(is.numeric), -elevational_range,-positive, -negative, -null,  -kipps_distance, -repro_age, -clutches_per_year, -lifespan, -eco_spec_index, -ends_with("_id"), -fstimb)
###################################################################################
#MAKE CORRELATION MATRIX
# M<-cor(trait.pca[,-1])
# res1 <- cor.mtest(trait.pca[,-1], conf.level = 0.95)
# ord <- corrMatOrder(M, order = "AOE")
# 
# png(here("Figures", "PCA","correlation analysis", "corrplot_output.png"), width = 2000, height = 2000, res = 150)
# corrplot(M, p.mat = res1$p, method="circle",type="upper",insig = "p-value", sig.level=-1)
# #so yes, some are correlated! onwards
# 
# png(here("Figures", "PCA", "correlation analysis", "scatterplot_output.png"), width = 3000, height = 3000, res = 150)
# par(oma = c(8, 8, 8, 8),      # huge outer margins for labels
#     mar = c(4, 4, 2, 2),
#     cex.labels = 2,           # bigger variable names
#     cex.axis = 1.2,
#     cex = 1.2)
# 
# pairs(trait.pca[,-1],
#       upper.panel = NULL,
#       pch = 20,
#       cex = 1.2)
# 
# 
# 
# ggpairs(
#   trait.pca[,-1],
#   upper = list(
#     continuous = wrap("cor", size = 4)   # top triangle = correlation coefficients
#   ),
#   lower = list(
#     continuous = wrap("points", alpha = 0.5, size = 1)   # bottom = scatterplots
#   ),
#   diag = list(
#     continuous = wrap("densityDiag")     # diagonal = density plots
#   )
# )

png(here("Figures", "PCA", "correlation analysis", "scatterplot_output.png"), width = 3000, height = 3000, res = 150)

ggpairs(
  trait.pca[,-1],
  upper = list(continuous = "cor"),  # correlation matrix
  lower = list(continuous = "points"),
  diag  = list(continuous = "densityDiag")
)

```


```{r}
trait.pca <- traits %>% 
  select(common_name, where(is.numeric), -elevational_range,-positive, -negative, -null,  -kipps_distance, -repro_age, -clutches_per_year, -lifespan, -eco_spec_index, -ends_with("_id"), -fstimb)

# trait_list <- trait.pca 
# trait_list <- trait_list[1, ]
# write.csv(trait_list, here("Data", "cleaned", "traits.for.pca.csv"), row.names = FALSE)


pca_result <- prcomp(trait.pca[,-1], scale = T) #scale the variables!
##
pca_result$rotation
pca_result$x

#autoplot(pca_result, data = climate, loadings=T,loadings.label=T, label = T)
biplot(pca_result)

pca_loadings <- as.data.frame(pca_result$rotation) %>% select(1:10)

summary(pca_result)
VE <- pca_result$sdev^2
PVE <- VE / sum(VE)
PVE <- round(PVE, 2)

pve_df <- data.frame(
  PC = 1:length(PVE),
  PVE = PVE
)

(scree <- ggplot(pve_df, aes(x = PC, y = PVE)) +
  geom_col() +
  xlab("Principal Component") +
  ylab("Proportion of Variance Explained") +
  ggtitle("Scree Plot") +
  ylim(0, 1) +
    theme_bw()+
  scale_x_continuous(breaks = 1:length(PVE)))
ggsave(here("Figures", "PCA", "scree plot.png"), plot = scree, width = 8, height =8)


##
pca_df <- as.data.frame(pca_result$x) %>%
  bind_cols(traits %>% select(common_name, positive, negative, null))
pca_df <- pca_df %>%
  mutate(fire_response = case_when(
    positive == 1 & negative == 0 & null == 0 ~ "Positive",
    positive == 0 & negative == 1 & null == 0 ~ "Negative",
    positive == 0 & negative == 0 & null == 1 ~ "Neutral",
    positive == 1 & negative == 1 & null == 0 ~ "Positive & Negative",
    positive == 1 & null == 1 & negative == 0 ~ "Positive & Neutral",
    negative == 1 & null == 1 & positive == 0 ~ "Negative & Neutral",
    TRUE ~ "None"
  ))

loadings <- as.data.frame(pca_result$rotation)  # columns = PCs, rows = traits
loadings$trait <- rownames(loadings)

# Optional: scale arrows for visualization
arrow_multiplier <- 5  # adjust to make arrows visible
loadings <- loadings %>%
  mutate(PC1_arrow = PC1 * arrow_multiplier,
         PC2_arrow = PC2 * arrow_multiplier)


(f1 <- ggplot(pca_df, aes(PC1, PC2, color = fire_response)) +
  geom_point(size = 2, alpha = 0.9) +
  theme_classic() +
  labs(
    title = "PCA of Bird Traits Colored by Fire Response",
    x = paste0("PC1 (", round(summary(pca_result)$importance[2,1]*100, 1), "%)"),
    y = paste0("PC2 (", round(summary(pca_result)$importance[2,2]*100, 1), "%)")
  ) +
   geom_segment(data = loadings,
               aes(x = 0, y = 0, xend = PC1_arrow, yend = PC2_arrow),
               arrow = arrow(length = unit(0.3, "cm")),
               color = "black") +
  geom_text(data = loadings,
            aes(x = PC1_arrow * 1.1, y = PC2_arrow * 1.1, label = trait),
            color = "black", size = 3, hjust = 0.5, vjust = 0.5) +
  scale_color_manual(values = c(
    "Positive" = "#005AB5",
    "Negative" = "#DC3220",
    "Neutral" = "grey",
    "Positive & Negative" = "#44AA99",
    "Positive & Neutral" = "#66a61e",
    "Negative & Neutral" = "#e6ab02"
  )))
(f1.1 <- ggplot(pca_df, aes(PC1, PC2, color = fire_response)) +
  geom_point(size = 2, alpha = 0.9) +
  theme_classic() +
  labs(
    title = "PCA of Bird Traits Colored by Fire Response",
    x = paste0("PC1 (", round(summary(pca_result)$importance[2,1]*100, 1), "%)"),
    y = paste0("PC2 (", round(summary(pca_result)$importance[2,2]*100, 1), "%)")
  ) +
   geom_segment(data = loadings,
               aes(x = 0, y = 0, xend = PC1_arrow, yend = PC2_arrow),
               arrow = arrow(length = unit(0.3, "cm")),
               color = "black") +
  geom_text(data = loadings,
            aes(x = PC1_arrow * 1.1, y = PC2_arrow * 1.1, label = trait),
            color = "black", size = 3, hjust = 0.5, vjust = 0.5) +
  scale_color_manual(values = c(
    "Positive" = "#005AB5",
    "Negative" = "#DC3220",
    "Neutral" = "grey",
    "Positive & Negative" = "#44AA99",
    "Positive & Neutral" = "#005AB5",
    "Negative & Neutral" = "#DC3220"
  )))
ggsave(here("Figures", "PCA", "all responses.png"), plot = f1, width = 10, height =8)

#If the colors are scattered randomly:
##Positive/negative response is not strongly associated with the traits included in the PCA.

# ggplot(pca_df, aes(x = PC1, y = PC2, color = Species, group = Species)) +
#       geom_point() + # Add points for individual data points
#       stat_ellipse(level = 0.95, type = "t") + # Add 95% confidence ellipses based on a t-distribution
#       labs(title = "PCA Plot with Confidence Ellipses",
#            x = "Principal Component 1",
#            y = "Principal Component 2") +
#       theme_minimal()
```

```{r}
arrow_multiplier <- 3
loadings <- loadings %>%
  mutate(PC1_arrow = PC1 * arrow_multiplier,
         PC2_arrow = PC2 * arrow_multiplier)

loadings.example <- loadings %>% 
  filter()

# Plot
(f2 <- ggplot() +
  geom_segment(data = loadings,
               aes(x = 0, y = 0, xend = PC1_arrow, yend = PC2_arrow),
               arrow = arrow(length = unit(0.3, "cm")),
               color = "black") +
  geom_text(data = loadings,
            aes(x = PC1_arrow * 1.1, y = PC2_arrow * 1.1, label = trait),
            color = "black", size = 3) +
  theme_bw() +
  labs(x = "PC1", y = "PC2"))
ggsave(here("Figures", "PCA", "pca loadings.png"), plot = f2, width = 10, height =8)


# Filter top 5 loadings for PC1 and PC2, excluding 'any_tree_nest'
top_PC1 <- loadings %>%
  filter(trait != "any_tree_nest") %>%
  slice_max(abs(PC1_arrow), n = 4)

top_PC2 <- loadings %>%
  filter(trait != "any_tree_nest") %>%
  slice_max(abs(PC2_arrow), n = 4)

# Combine top loadings
top_loadings <- bind_rows(top_PC1, top_PC2) %>%
  distinct()  # remove duplicates if a trait is top in both PCs

# Plot
(fex <- ggplot() +
  geom_segment(data = top_loadings,
               aes(x = 0, y = 0, xend = PC1_arrow, yend = PC2_arrow),
               arrow = arrow(length = unit(0.3, "cm")),
               color = "black") +
  geom_text_repel(data = top_loadings,
                  aes(x = PC1_arrow, y = PC2_arrow, label = trait),
                  color = "black",
                  size = 4,
                  segment.color = "grey50",    # line from label to point
                  segment.size = 0.5,
                  segment.alpha = 0.8,
                  nudge_x = 0,                 # optional small nudge
                  nudge_y = 0,
                  box.padding = 0.5,
                  point.padding = 0.5,
                  max.overlaps = 20) +
  theme_bw() +
  labs(x = "PC1", y = "PC2"))
ggsave(here("Figures", "PCA", "pca EXAMPLE.png"), plot = fex, width = 6, height =6)



```

```{r}
pca_scores <- as.data.frame(pca_result$x)
pca_scores$common_name <- traits$common_name
pca_scores$positive <- traits$positive
pca_scores$negative <- traits$negative
pca_scores$null <- traits$null

# 3️⃣ ggplot biplot colored by positive
(f3 <- ggplot(pca_scores, aes(x = PC1, y = PC2, color = factor(positive))) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_manual(values = c("0" = "grey", "1" = "#005AB5")) +
  theme_bw() +
  labs(color = "Positive Response", x = "PC1", y = "PC2",
       title = "PCA of Bird Traits Colored by Positive Fire Response"))
(f4 <- ggplot(pca_scores, aes(x = PC1, y = PC2, color = factor(negative))) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_manual(values = c("0" = "grey", "1" = "#DC3220")) +
  theme_bw() +
  labs(color = "Negative Response", x = "PC1", y = "PC2",
       title = "PCA of Bird Traits Colored by Negative Fire Response"))

(pos.neg <- f3/f4)
ggsave(here("Figures", "PCA", "pca by response.png"), plot = pos.neg, width = 8, height =10)


##
(f5 <- ggplot(pca_scores, aes(x = PC1, y = PC2, color = factor(positive))) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_manual(values = c("0" = "grey", "1" = "#005AB5")) +
  theme_bw() +
  geom_text(aes(label = common_name), size = 3) +
  labs(color = "Positive Response", x = "PC1", y = "PC2",
       title = "PCA of Bird Traits Colored by Positive Fire Response"))
(f6 <- ggplot(pca_scores, aes(x = PC1, y = PC2, color = factor(negative))) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_manual(values = c("0" = "grey", "1" = "#DC3220")) +
    geom_text(aes(label = common_name), size = 3) +
  theme_bw() +
  labs(color = "Negative Response", x = "PC1", y = "PC2",
       title = "PCA of Bird Traits Colored by Negative Fire Response"))


(pos.neg.spp.names <- f5/f6)
ggsave(here("Figures", "PCA", "pca by response spp names.png"), plot = pos.neg.spp.names, width = 20, height =20, dpi = 300)

```


```{r}
# Extract loadings
loadings <- as.data.frame(pca_result$rotation)

# Add trait names as a column
loadings$trait <- rownames(loadings)

loadings_long <- loadings %>%
  pivot_longer(cols = starts_with("PC"), 
               names_to = "PC", 
               values_to = "loading") %>%
  filter(PC %in% paste0("PC", 1:2))   # keep only PCs 1-5

(loadings.bar <- 
  ggplot(loadings_long, aes(x = trait, y = loading, fill = PC)) +
  geom_col(position = "dodge") +   # side-by-side bars per PC
  coord_flip() +                   # horizontal bars
  labs(x = "Trait", y = "Loading") +
  theme_bw(base_size = 14) +
  scale_y_continuous(limits = c(-0.5, 0.5), breaks = seq(-0.5, 0.5, 0.25)) +  # set range and breaks
  scale_fill_manual(values = c("#a1d99b", "#688c64")) +  # set all bars to green
  facet_wrap(~PC)+
    theme(legend.position = "none"))

ggsave(here("Figures", "PCA", "loadings bar plot.png"), plot = loadings.bar, width = 6, height =5, dpi = 300)

```

```{r}
# traits <- read.csv(here("Data", "cleaned", "BC_bird_traits.csv"))
# traits$habitat_density[traits$habitat_density == "dense"] <- 1
# traits$habitat_density[traits$habitat_density == "semiopen"] <- 2
# traits$habitat_density[traits$habitat_density == "open"] <- 3
# traits <- traits %>% 
#    mutate(cavity_nester = ifelse(nest_category == "cavity", 1, 0),
#           ground_nester = ifelse(most_common_nesting_location == "ground", 1, 0),
#           shrub_nester = ifelse(most_common_nesting_location == "shrub", 1, 0))
bc <- traits2 %>%  
  select(common_name, where(is.numeric), -elevational_range,  -kipps_distance, -repro_age, -clutches_per_year, -lifespan, -eco_spec_index, -ends_with("_id"), -fstimb) %>% 
  # mutate(null = NA,
  #        positive = NA,
  #        negative = NA) %>% 
  drop_na()  
#plotting all spp on this pca:
bc_scaled <- scale(bc[,-1],
                    center = (pca_result$center),
                    scale = (pca_result$scale))

bc_scores <- as.data.frame(as.matrix(bc_scaled) %*% pca_result$rotation)

```


```{r}
#bc_scores$common_name <- bc$common_name
pca_scores <- pca_scores %>%
  mutate(fire_response = case_when(
    positive == 1 & negative == 1 ~ "mixed",
    positive == 1 & negative == 0 ~ "positive",
    negative == 1 & positive == 0 ~ "negative",
    TRUE ~ "neutral"
  ))
centers <- data.frame(
  group = c("positive", "mixed", "negative", "neutral"),
  PC1 = c(
    mean(pca_scores$PC1[pca_scores$positive == 1 & pca_scores$negative != 1], na.rm = TRUE),
    mean(pca_scores$PC1[pca_scores$positive == 1 & pca_scores$negative == 1], na.rm = TRUE),
    mean(pca_scores$PC1[pca_scores$negative == 1 & pca_scores$positive != 1], na.rm = TRUE),
    mean(pca_scores$PC1[pca_scores$null == 1 & pca_scores$positive == 0 & pca_scores$negative == 0], na.rm = TRUE)),
  PC2 = c(
    mean(pca_scores$PC2[pca_scores$positive == 1 & pca_scores$negative != 1], na.rm = TRUE),
    mean(pca_scores$PC2[pca_scores$positive == 1 & pca_scores$negative == 1], na.rm = TRUE),
    mean(pca_scores$PC2[pca_scores$negative == 1 & pca_scores$positive != 1], na.rm = TRUE),
    mean(pca_scores$PC2[pca_scores$null == 1 & pca_scores$positive == 0 & pca_scores$negative == 0], na.rm = TRUE)
  )
)

# Basic PCA scatterplot
(all <- ggplot(bc_scores, aes(x = PC1, y = PC2)) +  # V1 and V2 are PC1 and PC2
  geom_point(size = 3, alpha = 0.7) +
  labs(x = "PC1", y = "PC2", title = "All BC Birds Plotted on PCA Axes") +
  theme_bw())
ggsave(here("Figures", "PCA", "all spp.png"), plot = all, width = 6, height =6, dpi = 300)


# Original PCA points
orig_df <- pca_scores
orig_df$type <- "PCA"

# New projected points
new_df <- bc_scores
new_df$common_name <- bc$common_name[complete.cases(bc)]


# Combine
plot_df <- left_join(new_df,orig_df)

# Overlay plot
(p1 <- ggplot(orig_df, aes(x = PC1, y = PC2)) +
    geom_point(data = new_df, aes( x = PC1, y = PC2), shape = 1, color = "grey", alpha = 0.85, size = 2)+
  geom_point(aes(color = factor(positive)), size = 2.5, alpha = 0.8) +
  scale_color_manual(values = c("0" = "grey50", "1" = "#005AB5")) +
  theme_bw() +
  labs(x = "PC1", y = "PC2", color = "Positive Response", shape = "Point Type"))

(n1 <- ggplot(orig_df, aes(x = PC1, y = PC2)) +
    geom_point(data = new_df, aes( x = PC1, y = PC2), shape = 1, color = "grey", alpha = 0.85, size = 2)+
  geom_point(aes(color = factor(negative)), size = 2.5, alpha = 0.8) +
  scale_color_manual(values = c("0" = "grey50", "1" = "#D55E00")) +
  theme_bw() +
  labs(x = "PC1", y = "PC2", color = "Negative Response", shape = "Point Type"))

orig_df <- orig_df %>% 
  mutate(
    mixed = rowSums(data.frame(positive, negative), na.rm = TRUE),
    mixed = ifelse(mixed == 2|mixed == 3, 1, 0))

(m1 <- ggplot(orig_df, aes(x = PC1, y = PC2)) +
    geom_point(data = new_df, aes( x = PC1, y = PC2), shape = 1, color = "grey", alpha = 0.85, size = 2)+
  geom_point(aes(color = factor(mixed)), size = 2.5, alpha = 0.8) +
  scale_color_manual(values = c("0" = "grey50", "1" = "darkmagenta")) +
  theme_bw() +
  labs(x = "PC1", y = "PC2", color = "Mixed Response", shape = "Point Type"))

(pcas <- p1/n1/m1 + plot_annotation(tag_levels = 'A')
)
ggsave(here("Figures", "PCA", "PCA_all response patchwork.png"), plot = pcas, width = 8, height =12, dpi = 300)


(with.mix <- ggplot(pca_scores) +
  geom_point(data = new_df, aes( x = PC1, y = PC2), shape = 1, color = "grey", alpha = 0.85, size = 2)+
  geom_point(aes(PC1, PC2), color = "grey80") +
  geom_point(data = subset(pca_scores, positive == 1),
             aes(PC1, PC2), color = "#0072B2", size = 3, alpha = 0.8) +
  geom_point(data = subset(pca_scores, negative == 1),
             aes(PC1, PC2), color = "#D55E00", size = 3, alpha = 0.8) +
  geom_point(data = subset(pca_scores, negative == 1& positive == 1),
             aes(PC1, PC2), color = "darkmagenta", size = 3, alpha = 0.8)+
  geom_point(data = centers, aes(PC1, PC2), size = 5, color = "white", shape = 16) +
  geom_point(data = centers, aes(PC1, PC2, color = group), size = 7, shape = 8) +
  scale_color_manual(values = c("darkmagenta", "#D55E00","grey80","#0072B2")) +
  theme_bw() +
  theme(legend.position="bottom")+
  labs(color = "Fire Response",
       title = ""))
ggsave(here("Figures", "PCA", "PCA_centers.png"), plot = with.mix, width = 8, height =8, dpi = 300)

(wo.mix <- ggplot(pca_scores) +
  geom_point(data = new_df, aes( x = PC1, y = PC2), shape = 1, color = "grey", alpha = 0.85, size = 2)+
  geom_point(aes(PC1, PC2), color = "grey80") +
  geom_point(data = subset(pca_scores, positive == 1),
             aes(PC1, PC2), color = "#0072B2", size = 3, alpha = 0.8) +
  geom_point(data = subset(pca_scores, negative == 1),
             aes(PC1, PC2), color = "#D55E00", size = 3, alpha = 0.8) +
  geom_point(data = centers, aes(PC1, PC2), size = 5, color = "white", shape = 16) +
  geom_point(data = centers, aes(PC1, PC2, color = group), size = 5, shape = 8) +
  scale_color_manual(values = c("darkmagenta", "#D55E00","grey80","#0072B2")) +
  theme_bw() +
  labs(color = "Response type",
       title = "Locations of positive vs negative responses in PCA space"))

t.test(
  pca_scores$PC1[pca_scores$positive == 1&pca_scores$negative != 1],
  pca_scores$PC1[pca_scores$negative == 1&pca_scores$positive != 1]
)

# PC2
t.test(
  pca_scores$PC2[pca_scores$positive == 1&pca_scores$negative != 1],
  pca_scores$PC2[pca_scores$negative == 1&pca_scores$positive != 1]
)

anova_pc1 <- aov(PC1 ~ fire_response, data = pca_scores)
summary(anova_pc1)

# Post-hoc (Tukey) test for pairwise comparisons
TukeyHSD(anova_pc1)

anova_pc2 <- aov(PC2~ fire_response, data = pca_scores)
summary(anova_pc2)

# Post-hoc (Tukey) test for pairwise comparisons
TukeyHSD(anova_pc2)

##
arrow_multiplier <- 3
loadings <- loadings %>%
  mutate(PC1_arrow = PC1 * arrow_multiplier,
         PC2_arrow = PC2 * arrow_multiplier)

```
> summary(anova_pc1)
               Df Sum Sq Mean Sq F value   Pr(>F)    
fire_response   3  191.6   63.86   15.17 9.96e-09 ***
Residuals     159  669.5    4.21                     
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
> # Post-hoc (Tukey) test for pairwise comparisons
> TukeyHSD(anova_pc1)
  Tukey multiple comparisons of means
    95% family-wise confidence level

Fit: aov(formula = PC1 ~ fire_response, data = pca_scores)

$fire_response
                        diff        lwr        upr     p adj
negative-mixed    -0.3038765 -1.6120795  1.0043265 0.9309546
neutral-mixed     -3.1731860 -4.5979908 -1.7483812 0.0000002
positive-mixed    -1.4907550 -2.7133738 -0.2681362 0.0099228
neutral-negative  -2.8693095 -4.1495019 -1.5891170 0.0000002
positive-negative -1.1868785 -2.2374120 -0.1363450 0.0199196
positive-neutral   1.6824309  0.4898312  2.8750307 0.0019091

> summary(anova_pc2)
               Df Sum Sq Mean Sq F value Pr(>F)  
fire_response   3   25.5   8.499   2.879 0.0378 *
Residuals     159  469.4   2.952                 
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
> # Post-hoc (Tukey) test for pairwise comparisons
> TukeyHSD(anova_pc2)
  Tukey multiple comparisons of means
    95% family-wise confidence level

Fit: aov(formula = PC2 ~ fire_response, data = pca_scores)

$fire_response
                        diff        lwr         upr     p adj
negative-mixed    -0.4811402 -1.5765785  0.61429805 0.6651232
neutral-mixed     -1.2498137 -2.4428898 -0.05673767 0.0361447
positive-mixed    -0.8522038 -1.8759771  0.17156957 0.1386786
neutral-negative  -0.7686735 -1.8406568  0.30330978 0.2487872
positive-negative -0.3710636 -1.2507394  0.50861227 0.6929792
positive-neutral   0.3976099 -0.6010267  1.39624654 0.7299017

```{r}

(with.mix <- ggplot() +
   geom_point(data = new_df, aes(PC1, PC2),
             shape = 1, color = "grey", alpha = 0.85, size = 2) +
  geom_point(data = pca_scores,
             aes(PC1, PC2, color = fire_response),
             size = 3, alpha = 0.8) +
  geom_point(data = centers, aes(PC1, PC2, color = group),
             size = 7, shape = 8) +
 
  scale_color_manual(
    values = c(
      "positive" = "#0072B2",
      "negative" = "#D55E00",
      "mixed" = "darkmagenta",
      "neutral" = "grey50"
    )
  ) +
  theme_bw() +
     scale_x_continuous(breaks = seq(floor(min(pca_scores$PC1)), 
                                  ceiling(max(pca_scores$PC1)), 
                                  by = 2))+
  theme(legend.position = "bottom") +
  labs(color = "Fire Response",
       title = "",x = paste0("PC1 (", round(summary(pca_result)$importance[2,1]*100, 1), "%)"),
    y = paste0("PC2 (", round(summary(pca_result)$importance[2,2]*100, 1), "%)")))

(f2 <- ggplot() +
  geom_segment(data = loadings,
               aes(x = 0, y = 0, xend = PC1_arrow, yend = PC2_arrow),
               arrow = arrow(length = unit(0.3, "cm")),
               color = "black") +
  geom_text_repel(data = loadings,
                  aes(x = PC1_arrow, y = PC2_arrow, label = trait),
                  color = "black",
                  size = 4,
                  segment.color = "grey50",    # line from label to point
                  segment.size = 0.5,
                  segment.alpha = 0.8,
                  nudge_x = 0,                 # optional small nudge
                  nudge_y = 0,
                  box.padding = 0.5,
                  point.padding = 0.5,
                  max.overlaps = 20) +
  geom_point(data = centers, aes(PC1, PC2, color = group), size = 6, shape = 8) +
  theme_bw() +
    theme(legend.position = "none")+
  scale_color_manual(values = c("darkmagenta", "#D55E00","grey50","#0072B2")) +
  labs(x = paste0("PC1 (", round(summary(pca_result)$importance[2,1]*100, 1), "%)"),
    y = paste0("PC2 (", round(summary(pca_result)$importance[2,2]*100, 1), "%)")))

(end <- with.mix+f2+  
  plot_layout(guides = "collect") +
  plot_annotation(theme = theme(legend.position = "bottom"), tag_levels = "A"))
ggsave(here("Figures", "PCA", "PCA_final.png"), plot = end, width = 18, height =8, dpi = 300)

```

