---
title: "PCA"
author: "JL"
date: "`r Sys.Date()`"
output: html_document
---
```{r}
library(here)
library(tidyverse)
library(patchwork)
library(ggrepel)
```

```{r}
traits <- read.csv(here("Data", "cleaned", "BC_bird_traits.csv"))
traits$habitat_density[traits$habitat_density == "dense"] <- 1
traits$habitat_density[traits$habitat_density == "semiopen"] <- 2
traits$habitat_density[traits$habitat_density == "open"] <- 3
traits <- traits %>% 
   mutate(cavity_nester = ifelse(nest_category == "cavity", 1, 0),
          ground_nester = ifelse(most_common_nesting_location == "ground", 1, 0),
          shrub_nester = ifelse(most_common_nesting_location == "shrub", 1, 0)) %>%
    mutate(habitat_density = as.numeric(habitat_density)) 
# old_growth <-  c("American Three-toed Woodpecker", "Bay-breasted Warbler", "Black-capped Chickadee", "Boreal Chickadee", "Boreal Owl", "Brown Creeper", "Canada Warbler", "Chestnut-backed Chickadee", "Dusky Grouse", "Golden-crowned Kinglet", "Great Horned Owl", "Hairy Woodpecker", "Hammond's Flycatcher", "Hermit Thrush", "Lewis's Woodpecker", "American Goshawk", "Northern Waterthrush", "Western Flycatcher", "Pileated Woodpecker", "Red-breasted Nuthatch", "Red-breasted Sapsucker", "Ruby-crowned Kinglet", "Spruce Grouse", "Swainson's Thrush", "Townsend's Warbler", "Varied Thrush", "Vaux's Swift", "White-breasted Nuthatch")
# missing_names <- old_growth[!(old_growth %in% traits$common_name)] #do all these names appear in the BC list?


##
positive <- c("Alder Flycatcher", "Canada Goose", "Swamp Sparrow", "Red-winged Blackbird", "Common Yellowthroat", "Western Bluebird", "Steller's Jay", "Hairy Woodpecker", "American Robin", "Northern Flicker", "Olive-sided Flycatcher", "Mountain Bluebird", "American Three-toed Woodpecker", "Black-backed Woodpecker", "Western Wood-Pewee", "Northern House Wren", "Lazuli Bunting", "Cassin's Finch", "Pine Siskin","Black-backed Woodpecker", "Lewis's Woodpecker", "Dark-eyed Junco", "Townsend's Solitaire", "Acorn Woodpecker", "Tree Swallow", "American Kestrel", "Northern Hawk Owl", "White-throated Sparrow", "Chipping Sparrow", "White-crowned Sparrow", "Brown Creeper", "Tennessee Warbler", "Chestnut-sided Warbler", "Brewer's Sparrow", "Rock Wren", "Hermit Thrush", "White-breasted Nuthatch", "American Crow", "American Goldfinch", "Baltimore Oriole", "Black-billed Magpie", "Brown-headed Cowbird", "Clay-colored Sparrow", "Eastern Kingbird", "Nelson's Sparrow", "Savannah Sparrow", "Song Sparrow", "Yellow-bellied Sapsucker", "Canada Jay", "Lincoln's Sparrow", "MacGillivray's Warbler", "Spotted Towhee", "Turkey Vulture", "Brewer's Blackbird", "Say's Phoebe", "Warbling Vireo", "Veery", "Song Sparrow", "Red-breasted Nuthatch", "Mountain Chickadee", "Redpoll", "Western Screech-Owl", "Northern Pygmy-Owl", "Mountain Quail", "Black-capped Chickadee", "Downy Woodpecker", "Pacific Wren", "Connecticut Warbler", "Yellow-rumped Warbler", "Cedar Waxwing", "Eastern Bluebird", "Broad-tailed Hummingbird", "Red-headed Woodpecker", "Eastern Kingbird", "Evening Grosbeak", "Prairie Warbler", "Palm Warbler", "Field Sparrow", "Black-headed Grosbeak", "Black-throated Gray Warbler", "Western Tanager",  "Dusky Flycatcher", "Hermit Warbler")
missing_names <- positive[!(positive %in% traits$common_name)] #do all these names appear in the BC list?

negative <- c("American Black Duck", "Common Goldeneye", "Western Wood-Pewee", "Pygmy Nuthatch", "Cassin's Vireo", "Dusky Flycatcher", "Ruby-crowned Kinglet", "Golden-crowned Kinglet", "Swainson's Thrush", "Yellow-rumped Warbler", "Townsend's Warbler", "Swainson's Thrush", "Hermit Warbler", "Red-breasted Nuthatch", "Brown Creeper", "Chestnut-backed Chickadee", "Black-throated Gray Warbler", "Winter Wren", "Hutton's Vireo", "Western Tanager", "Steller's Jay", "Wilson's Warbler", "Common Raven", "Sooty Grouse", "Northern Waterthrush", "Blue-headed Vireo", "Red-eyed Vireo", "Ovenbird", "MacGillivray's Warbler", "Hermit Thrush", "Nashville Warbler", "Hermit Warbler", "Warbling Vireo", "Mountain Chickadee", "Dark-eyed Junco", "Wilson's Warbler", "Philadelphia Vireo", "Canada Jay", "Pine Grosbeak", "Red Crossbill", "Yellow Warbler", "Bay-breasted Warbler", "Cape May Warbler", "Purple Finch",  "Boreal Chickadee", "Pine Siskin", "White-winged Crossbill", "Alder Flycatcher", "Common Yellowthroat", "Orange-crowned Warbler", "Ruffed Grouse", "Swamp Sparrow", "Yellow-rumped Warbler", "Pinyon Jay", "Greater Sage-Grouse", "Red-naped Sapsucker", "Barred Owl", "Northern Saw-whet Owl", "Great Horned Owl", "Vaux's Swift", "Red-breasted Sapsucker", "Pileated Woodpecker", "Red-breasted Nuthatch", "Western Tanager", "Red Crossbill", "Acorn Woodpecker",  "Williamson's Sapsucker", "Acadian Flycatcher",  "Ash-throated Flycatcher", "Great Crested Flycatcher", "Blue-gray Gnatcatcher", "Wood Thrush", "Black-and-white Warbler", "Hooded Warbler", "Northern Cardinal", "Black-headed Grosbeak", "Pacific Wren")
missing_names <- negative[!(negative %in% traits$common_name)] #do all these names appear in the BC list?

null <- c("American Wigeon", "Green-winged Teal", "Mallard", "Northern Pintail", "Northern Shoveler", "Bufflehead", "Canvasback", "Barrow's Goldeneye", "Common Goldeneye", "Red-breasted Merganser", "Common Merganser","Ring-necked Duck", "Greater Scaup", "Lesser Scaup", "Surf Scoter", "White-winged Scoter", "Mallard", "Ring-necked Duck", "Common Loon", "Mourning Dove", "Common Nighthawk", "Cassin's Finch","Pine Siskin", "Chipping Sparrow", "Dark-eyed Junco", "American Robin", "Townsend's Solitaire", "Hammond's Flycatcher", "Clark's Nutcracker", "Red-naped Sapsucker", "Western Tanager", "White-breasted Nuthatch", "Evening Grosbeak", "Pygmy Nuthatch", "Yellow-rumped Warbler", "Williamson's Sapsucker", "Red Crossbill", "Ruby-crowned Kinglet", "Pacific Wren", "Brewer's Blackbird", "Mountain Chickadee", "Nashville Warbler", "Swainson's Thrush",  "House Finch", "Red Crossbill", "American Goldfinch", "American Kestrel", "Red-breasted Sapsucker", "Hairy Woodpecker", "Olive-sided Flycatcher", "Cassin's Vireo", "Wilson's Warbler")
missing_names <- null[!(null %in% traits$common_name)] #do all these names appear in the BC list?

traits$positive <- ifelse(traits$common_name %in% positive, 1, NA)
traits$negative <- ifelse(traits$common_name %in% negative, 1, NA)
traits$null <- ifelse(traits$common_name %in% null, 1, NA)


traits <- traits %>%
    filter(!if_all(c(positive, negative, null), is.na)) %>% 
    mutate(across(c(positive, negative, null), ~replace_na(.x, 0)))



```

```{r}
trait.pca <- traits %>% select(common_name, where(is.numeric), -elevational_range,-positive, -negative, -null, -nest_tree_use, -kipps_distance, -repro_age, -clutches_per_year)

# trait_list <- trait.pca 
# trait_list <- trait_list[1, ]
# write.csv(trait_list, here("Data", "cleaned", "traits.for.pca.csv"), row.names = FALSE)


pca_result <- prcomp(trait.pca[,-1], scale = T) #scale the variables!
##
pca_result$rotation
pca_result$x

#autoplot(pca_result, data = climate, loadings=T,loadings.label=T, label = T)
biplot(pca_result)

pca_loadings <- as.data.frame(pca_result$rotation) %>% select(1:10)

summary(pca_result)
VE <- pca_result$sdev^2
PVE <- VE / sum(VE)
PVE <- round(PVE, 2)

pve_df <- data.frame(
  PC = 1:length(PVE),
  PVE = PVE
)

(scree <- ggplot(pve_df, aes(x = PC, y = PVE)) +
  geom_col() +
  xlab("Principal Component") +
  ylab("Proportion of Variance Explained") +
  ggtitle("Scree Plot") +
  ylim(0, 1) +
    theme_bw()+
  scale_x_continuous(breaks = 1:length(PVE)))
ggsave(here("Figures", "PCA", "scree plot.png"), plot = scree, width = 8, height =8)


##
pca_df <- as.data.frame(pca_result$x) %>%
  bind_cols(traits %>% select(common_name, positive, negative, null))
pca_df <- pca_df %>%
  mutate(fire_response = case_when(
    positive == 1 & negative == 0 & null == 0 ~ "Positive",
    positive == 0 & negative == 1 & null == 0 ~ "Negative",
    positive == 0 & negative == 0 & null == 1 ~ "Neutral",
    positive == 1 & negative == 1 & null == 0 ~ "Positive & Negative",
    positive == 1 & null == 1 & negative == 0 ~ "Positive & Neutral",
    negative == 1 & null == 1 & positive == 0 ~ "Negative & Neutral",
    TRUE ~ "None"
  ))

loadings <- as.data.frame(pca_result$rotation)  # columns = PCs, rows = traits
loadings$trait <- rownames(loadings)

# Optional: scale arrows for visualization
arrow_multiplier <- 5  # adjust to make arrows visible
loadings <- loadings %>%
  mutate(PC1_arrow = PC1 * arrow_multiplier,
         PC2_arrow = PC2 * arrow_multiplier)


(f1 <- ggplot(pca_df, aes(PC1, PC2, color = fire_response)) +
  geom_point(size = 2, alpha = 0.9) +
  theme_classic() +
  labs(
    title = "PCA of Bird Traits Colored by Fire Response",
    x = paste0("PC1 (", round(summary(pca_result)$importance[2,1]*100, 1), "%)"),
    y = paste0("PC2 (", round(summary(pca_result)$importance[2,2]*100, 1), "%)")
  ) +
   geom_segment(data = loadings,
               aes(x = 0, y = 0, xend = PC1_arrow, yend = PC2_arrow),
               arrow = arrow(length = unit(0.3, "cm")),
               color = "black") +
  geom_text(data = loadings,
            aes(x = PC1_arrow * 1.1, y = PC2_arrow * 1.1, label = trait),
            color = "black", size = 3, hjust = 0.5, vjust = 0.5) +
  scale_color_manual(values = c(
    "Positive" = "#005AB5",
    "Negative" = "#DC3220",
    "Neutral" = "grey",
    "Positive & Negative" = "#44AA99",
    "Positive & Neutral" = "#66a61e",
    "Negative & Neutral" = "#e6ab02"
  )))
(f1.1 <- ggplot(pca_df, aes(PC1, PC2, color = fire_response)) +
  geom_point(size = 2, alpha = 0.9) +
  theme_classic() +
  labs(
    title = "PCA of Bird Traits Colored by Fire Response",
    x = paste0("PC1 (", round(summary(pca_result)$importance[2,1]*100, 1), "%)"),
    y = paste0("PC2 (", round(summary(pca_result)$importance[2,2]*100, 1), "%)")
  ) +
   geom_segment(data = loadings,
               aes(x = 0, y = 0, xend = PC1_arrow, yend = PC2_arrow),
               arrow = arrow(length = unit(0.3, "cm")),
               color = "black") +
  geom_text(data = loadings,
            aes(x = PC1_arrow * 1.1, y = PC2_arrow * 1.1, label = trait),
            color = "black", size = 3, hjust = 0.5, vjust = 0.5) +
  scale_color_manual(values = c(
    "Positive" = "#005AB5",
    "Negative" = "#DC3220",
    "Neutral" = "grey",
    "Positive & Negative" = "#44AA99",
    "Positive & Neutral" = "#005AB5",
    "Negative & Neutral" = "#DC3220"
  )))
ggsave(here("Figures", "PCA", "all responses.png"), plot = f1, width = 10, height =8)

#If the colors are scattered randomly:
##Positive/negative response is not strongly associated with the traits included in the PCA.

# ggplot(pca_df, aes(x = PC1, y = PC2, color = Species, group = Species)) +
#       geom_point() + # Add points for individual data points
#       stat_ellipse(level = 0.95, type = "t") + # Add 95% confidence ellipses based on a t-distribution
#       labs(title = "PCA Plot with Confidence Ellipses",
#            x = "Principal Component 1",
#            y = "Principal Component 2") +
#       theme_minimal()
```

```{r}
arrow_multiplier <- 3
loadings <- loadings %>%
  mutate(PC1_arrow = PC1 * arrow_multiplier,
         PC2_arrow = PC2 * arrow_multiplier)

# Plot
(f2 <- ggplot() +
  geom_segment(data = loadings,
               aes(x = 0, y = 0, xend = PC1_arrow, yend = PC2_arrow),
               arrow = arrow(length = unit(0.3, "cm")),
               color = "black") +
  geom_text(data = loadings,
            aes(x = PC1_arrow * 1.1, y = PC2_arrow * 1.1, label = trait),
            color = "black", size = 3) +
  theme_bw() +
  labs(x = "PC1", y = "PC2"))
ggsave(here("Figures", "PCA", "pca loadings.png"), plot = f2, width = 10, height =8)

```

```{r}
pca_scores <- as.data.frame(pca_result$x)
pca_scores$common_name <- traits$common_name
pca_scores$positive <- traits$positive
pca_scores$negative <- traits$negative
pca_scores$null <- traits$null

# 3️⃣ ggplot biplot colored by positive
(f3 <- ggplot(pca_scores, aes(x = PC1, y = PC2, color = factor(positive))) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_manual(values = c("0" = "grey", "1" = "#005AB5")) +
  theme_bw() +
  labs(color = "Positive Response", x = "PC1", y = "PC2",
       title = "PCA of Bird Traits Colored by Positive Fire Response"))
(f4 <- ggplot(pca_scores, aes(x = PC1, y = PC2, color = factor(negative))) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_manual(values = c("0" = "grey", "1" = "#DC3220")) +
  theme_bw() +
  labs(color = "Negative Response", x = "PC1", y = "PC2",
       title = "PCA of Bird Traits Colored by Negative Fire Response"))

(pos.neg <- f3/f4)
ggsave(here("Figures", "PCA", "pca by response.png"), plot = pos.neg, width = 8, height =10)


##
(f5 <- ggplot(pca_scores, aes(x = PC1, y = PC2, color = factor(positive))) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_manual(values = c("0" = "grey", "1" = "#005AB5")) +
  theme_bw() +
  geom_text(aes(label = common_name), size = 3) +
  labs(color = "Positive Response", x = "PC1", y = "PC2",
       title = "PCA of Bird Traits Colored by Positive Fire Response"))
(f6 <- ggplot(pca_scores, aes(x = PC1, y = PC2, color = factor(negative))) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_manual(values = c("0" = "grey", "1" = "#DC3220")) +
    geom_text(aes(label = common_name), size = 3) +
  theme_bw() +
  labs(color = "Negative Response", x = "PC1", y = "PC2",
       title = "PCA of Bird Traits Colored by Negative Fire Response"))

(pos.neg.spp.names <- f5/f6)
ggsave(here("Figures", "PCA", "pca by response spp names.png"), plot = pos.neg.spp.names, width = 20, height =20, dpi = 300)

```


```{r}
# Extract loadings
loadings <- as.data.frame(pca_result$rotation)

# Add trait names as a column
loadings$trait <- rownames(loadings)

loadings_long <- loadings %>%
  pivot_longer(cols = starts_with("PC"), 
               names_to = "PC", 
               values_to = "loading") %>%
  filter(PC %in% paste0("PC", 1:5))   # keep only PCs 1-5

(loadings.bar <- ggplot(loadings_long, aes(x = trait, y = loading, fill = PC)) +
  geom_col(position = "dodge") +   # side-by-side bars per PC
  coord_flip() +                   # horizontal bars
  labs(x = "Trait", y = "Loading", 
       title = "PCA Loadings per Trait") +
  theme_bw(base_size = 14) +
  scale_fill_brewer(palette = "Set2")+facet_wrap(~PC)+
    theme(legend.position = "none"))

ggsave(here("Figures", "PCA", "loadings bar plot.png"), plot = loadings.bar, width = 15, height =15, dpi = 300)

```

```{r}
# traits <- read.csv(here("Data", "cleaned", "BC_bird_traits.csv"))
# traits$habitat_density[traits$habitat_density == "dense"] <- 1
# traits$habitat_density[traits$habitat_density == "semiopen"] <- 2
# traits$habitat_density[traits$habitat_density == "open"] <- 3
# traits <- traits %>% 
#    mutate(cavity_nester = ifelse(nest_category == "cavity", 1, 0),
#           ground_nester = ifelse(most_common_nesting_location == "ground", 1, 0),
#           shrub_nester = ifelse(most_common_nesting_location == "shrub", 1, 0))
bc <- traits %>% select(common_name, where(is.numeric), -elevational_range, -nest_tree_use, -kipps_distance, -repro_age, -clutches_per_year, -null, -negative, -positive) %>% 
  # mutate(null = NA,
  #        positive = NA,
  #        negative = NA) %>% 
  filter(if_all(1:22, ~ !is.na(.)))

#plotting all spp on this pca:
bc_scaled <- scale(bc[,-1],
                    center = (pca_result$center),
                    scale = (pca_result$scale))

bc_scores <- as.data.frame(as.matrix(bc_scaled) %*% pca_result$rotation)

```


```{r}
bc_scores$common_name <- bc$common_name
pca_scores <- pca_scores %>%
  mutate(fire_response = case_when(
    positive == 1 & negative == 1 ~ "mixed",
    positive == 1 & negative == 0 ~ "positive",
    negative == 1 & positive == 0 ~ "negative",
    TRUE ~ "neutral"
  ))
centers <- data.frame(
  group = c("positive", "mixed", "negative", "neutral"),
  PC1 = c(
    mean(pca_scores$PC1[pca_scores$positive == 1 & pca_scores$negative != 1], na.rm = TRUE),
    mean(pca_scores$PC1[pca_scores$positive == 1 & pca_scores$negative == 1], na.rm = TRUE),
    mean(pca_scores$PC1[pca_scores$negative == 1 & pca_scores$positive != 1], na.rm = TRUE),
    mean(pca_scores$PC1[pca_scores$null == 1 & pca_scores$positive == 0 & pca_scores$negative == 0], na.rm = TRUE)),
  PC2 = c(
    mean(pca_scores$PC2[pca_scores$positive == 1 & pca_scores$negative != 1], na.rm = TRUE),
    mean(pca_scores$PC2[pca_scores$positive == 1 & pca_scores$negative == 1], na.rm = TRUE),
    mean(pca_scores$PC2[pca_scores$negative == 1 & pca_scores$positive != 1], na.rm = TRUE),
    mean(pca_scores$PC2[pca_scores$null == 1 & pca_scores$positive == 0 & pca_scores$negative == 0], na.rm = TRUE)
  )
)

# Basic PCA scatterplot
(all <- ggplot(bc_scores, aes(x = PC1, y = PC2)) +  # V1 and V2 are PC1 and PC2
  geom_point(size = 3, alpha = 0.7) +
  labs(x = "PC1", y = "PC2", title = "All BC Birds Plotted on PCA Axes") +
  theme_bw())
ggsave(here("Figures", "PCA", "all spp.png"), plot = all, width = 15, height =15, dpi = 300)


# Original PCA points
orig_df <- pca_scores
orig_df$type <- "PCA"

# New projected points
new_df <- bc_scores
new_df$common_name <- bc$common_name[complete.cases(bc)]


# Combine
plot_df <- left_join(new_df,orig_df)

# Overlay plot
(p1 <- ggplot(orig_df, aes(x = PC1, y = PC2)) +
    geom_point(data = new_df, aes( x = PC1, y = PC2), shape = 1, color = "grey", alpha = 0.85, size = 2)+
  geom_point(aes(color = factor(positive)), size = 2.5, alpha = 0.8) +
  scale_color_manual(values = c("0" = "grey50", "1" = "#005AB5")) +
  theme_bw() +
  labs(x = "PC1", y = "PC2", color = "Positive Response", shape = "Point Type"))

(n1 <- ggplot(orig_df, aes(x = PC1, y = PC2)) +
    geom_point(data = new_df, aes( x = PC1, y = PC2), shape = 1, color = "grey", alpha = 0.85, size = 2)+
  geom_point(aes(color = factor(negative)), size = 2.5, alpha = 0.8) +
  scale_color_manual(values = c("0" = "grey50", "1" = "#D55E00")) +
  theme_bw() +
  labs(x = "PC1", y = "PC2", color = "Negative Response", shape = "Point Type"))

orig_df <- orig_df %>% 
  mutate(
    mixed = rowSums(data.frame(positive, negative), na.rm = TRUE),
    mixed = ifelse(mixed == 2|mixed == 3, 1, 0))

(m1 <- ggplot(orig_df, aes(x = PC1, y = PC2)) +
    geom_point(data = new_df, aes( x = PC1, y = PC2), shape = 1, color = "grey", alpha = 0.85, size = 2)+
  geom_point(aes(color = factor(mixed)), size = 2.5, alpha = 0.8) +
  scale_color_manual(values = c("0" = "grey50", "1" = "darkmagenta")) +
  theme_bw() +
  labs(x = "PC1", y = "PC2", color = "Mixed Response", shape = "Point Type"))

(pcas <- p1/n1/m1 + plot_annotation(tag_levels = 'A')
)
ggsave(here("Figures", "PCA", "PCA_all response patchwork.png"), plot = pcas, width = 8, height =12, dpi = 300)


(with.mix <- ggplot(pca_scores) +
  geom_point(data = new_df, aes( x = PC1, y = PC2), shape = 1, color = "grey", alpha = 0.85, size = 2)+
  geom_point(aes(PC1, PC2), color = "grey80") +
  geom_point(data = subset(pca_scores, positive == 1),
             aes(PC1, PC2), color = "#0072B2", size = 3, alpha = 0.8) +
  geom_point(data = subset(pca_scores, negative == 1),
             aes(PC1, PC2), color = "#D55E00", size = 3, alpha = 0.8) +
  geom_point(data = subset(pca_scores, negative == 1& positive == 1),
             aes(PC1, PC2), color = "darkmagenta", size = 3, alpha = 0.8)+
  geom_point(data = centers, aes(PC1, PC2), size = 5, color = "white", shape = 16) +
  geom_point(data = centers, aes(PC1, PC2, color = group), size = 7, shape = 8) +
  scale_color_manual(values = c("darkmagenta", "#D55E00","grey80","#0072B2")) +
  theme_bw() +
  theme(legend.position="bottom")+
  labs(color = "Fire Response",
       title = ""))
ggsave(here("Figures", "PCA", "PCA_centers.png"), plot = with.mix, width = 8, height =8, dpi = 300)

(wo.mix <- ggplot(pca_scores) +
  geom_point(data = new_df, aes( x = PC1, y = PC2), shape = 1, color = "grey", alpha = 0.85, size = 2)+
  geom_point(aes(PC1, PC2), color = "grey80") +
  geom_point(data = subset(pca_scores, positive == 1),
             aes(PC1, PC2), color = "#0072B2", size = 3, alpha = 0.8) +
  geom_point(data = subset(pca_scores, negative == 1),
             aes(PC1, PC2), color = "#D55E00", size = 3, alpha = 0.8) +
  geom_point(data = centers, aes(PC1, PC2), size = 5, color = "white", shape = 16) +
  geom_point(data = centers, aes(PC1, PC2, color = group), size = 5, shape = 8) +
  scale_color_manual(values = c("darkmagenta", "#D55E00","grey80","#0072B2")) +
  theme_bw() +
  labs(color = "Response type",
       title = "Locations of positive vs negative responses in PCA space"))

t.test(
  pca_scores$PC1[pca_scores$positive == 1&pca_scores$negative != 1],
  pca_scores$PC1[pca_scores$negative == 1&pca_scores$positive != 1]
)

# PC2
t.test(
  pca_scores$PC2[pca_scores$positive == 1&pca_scores$negative != 1],
  pca_scores$PC2[pca_scores$negative == 1&pca_scores$positive != 1]
)

anova_pc1 <- aov(PC1 ~ fire_response, data = pca_scores)
summary(anova_pc1)

# Post-hoc (Tukey) test for pairwise comparisons
TukeyHSD(anova_pc1)

anova_pc2 <- aov(PC2~ fire_response, data = pca_scores)
summary(anova_pc2)

# Post-hoc (Tukey) test for pairwise comparisons
TukeyHSD(anova_pc2)

##
arrow_multiplier <- 3
loadings <- loadings %>%
  mutate(PC1_arrow = PC1 * arrow_multiplier,
         PC2_arrow = PC2 * arrow_multiplier)

```

	Welch Two Sample t-test

data:  pca_scores$PC1[pca_scores$positive == 1 & pca_scores$negative != 1] and pca_scores$PC1[pca_scores$negative == 1 & pca_scores$positive != 1]
t = -3.2169, df = 95.089, p-value = 0.001772
alternative hypothesis: true difference in means is not equal to 0
95 percent confidence interval:
 -1.9603558 -0.4641274
sample estimates:
 mean of x  mean of y 
-0.2840188  0.9282228

	Welch Two Sample t-test

data:  pca_scores$PC2[pca_scores$positive == 1 & pca_scores$negative != 1] and pca_scores$PC2[pca_scores$negative == 1 & pca_scores$positive != 1]
t = 0.32073, df = 98.053, p-value = 0.7491
alternative hypothesis: true difference in means is not equal to 0
95 percent confidence interval:
 -0.5732841  0.7943165
sample estimates:
 mean of x  mean of y 
0.11210649 0.00159028 

```{r}

(with.mix <- ggplot() +
  geom_point(data = new_df, aes(PC1, PC2),
             shape = 1, color = "grey", alpha = 0.85, size = 2) +
  geom_point(data = pca_scores,
             aes(PC1, PC2, color = fire_response),
             size = 3, alpha = 0.8) +
  geom_point(data = centers, aes(PC1, PC2, color = group),
             size = 7, shape = 8) +
  scale_color_manual(
    values = c(
      "positive" = "#0072B2",
      "negative" = "#D55E00",
      "mixed" = "darkmagenta",
      "neutral" = "grey"
    )
  ) +
  theme_bw() +
     scale_x_continuous(breaks = seq(floor(min(pca_scores$PC1)), 
                                  ceiling(max(pca_scores$PC1)), 
                                  by = 2))+
  theme(legend.position = "bottom") +
  labs(color = "Fire Response",
       title = "",x = paste0("PC1 (", round(summary(pca_result)$importance[2,1]*100, 1), "%)"),
    y = paste0("PC2 (", round(summary(pca_result)$importance[2,2]*100, 1), "%)")))

(f2 <- ggplot() +
  geom_segment(data = loadings,
               aes(x = 0, y = 0, xend = PC1_arrow, yend = PC2_arrow),
               arrow = arrow(length = unit(0.3, "cm")),
               color = "black") +
  geom_text_repel(data = loadings,
                  aes(x = PC1_arrow, y = PC2_arrow, label = trait),
                  color = "black",
                  size = 4,
                  segment.color = "grey50",    # line from label to point
                  segment.size = 0.5,
                  segment.alpha = 0.8,
                  nudge_x = 0,                 # optional small nudge
                  nudge_y = 0,
                  box.padding = 0.5,
                  point.padding = 0.5,
                  max.overlaps = 20) +
  geom_point(data = centers, aes(PC1, PC2, color = group), size = 6, shape = 8) +
  theme_bw() +
    theme(legend.position = "none")+
  scale_color_manual(values = c("darkmagenta", "#D55E00","grey","#0072B2")) +
  labs(x = paste0("PC1 (", round(summary(pca_result)$importance[2,1]*100, 1), "%)"),
    y = paste0("PC2 (", round(summary(pca_result)$importance[2,2]*100, 1), "%)")))

(end <- with.mix+f2+  
  plot_layout(guides = "collect") +
  plot_annotation(theme = theme(legend.position = "bottom"), tag_levels = "A"))
ggsave(here("Figures", "PCA", "PCA_final.png"), plot = end, width = 18, height =8, dpi = 300)

```

