---
title: "pca_time"
author: "JL"
date: "`r Sys.Date()`"
output: html_document
---
```{r}
library(here)
library(tidyverse)
library(patchwork)
```

```{r SHORT TERM}
traits <- read.csv(here("Data", "cleaned", "BC_bird_traits.csv"))
traits$habitat_density[traits$habitat_density == "dense"] <- 1
traits$habitat_density[traits$habitat_density == "semiopen"] <- 2
traits$habitat_density[traits$habitat_density == "open"] <- 3
traits <- traits %>% 
    select(-any_tree_nest, -live_tree_nest) %>% 
    mutate(cavity_nester = ifelse(nest_category == "cavity", 1, 0),
   #        ground_nester = ifelse(most_common_nesting_location == "ground", 1, 0),
   #        shrub_nester = ifelse(most_common_nesting_location == "shrub", 1, 0)) %>%
    habitat_density = as.numeric(habitat_density))
traits2 <- traits

##
positive_short <- c("American Kestrel", "American Robin", "American Three-toed Woodpecker", "Black-backed Woodpecker", "Black-capped Chickadee", "Black-headed Grosbeak", "Black-throated Gray Warbler", "Brewer's Blackbird", "Brewer's Sparrow", "Brown Creeper", "Brown-headed Cowbird", "Canada Jay", "Cassin's Finch", "Cassin's Vireo", "Cedar Waxwing", "Chipping Sparrow", "Connecticut Warbler", "Dark-eyed Junco", "Downy Woodpecker", "Dusky Flycatcher","Fox Sparrow", "Hairy Woodpecker", "Hermit Thrush", "Northern House Wren", "Lazuli Bunting", "Least Flycatcher", "Lewis's Woodpecker", "Lincoln's Sparrow", "MacGillivray's Warbler", "Mountain Bluebird", "Mountain Chickadee", "Nashville Warbler", "Northern Flicker", "Northern Pygmy-Owl", "Olive-sided Flycatcher", "Pine Siskin", "Red-breasted Nuthatch", "Red-winged Blackbird", "Rock Wren", "Rufous Hummingbird", "Savannah Sparrow", "Say's Phoebe", "Song Sparrow", "Spotted Towhee", "Steller's Jay", "Tennessee Warbler", "Townsend's Solitaire", "Tree Swallow", "Turkey Vulture", "Veery", "Warbling Vireo", "Western Bluebird", "Western Screech-Owl", "Western Tanager", "Western Wood-Pewee", "White-breasted Nuthatch", "White-crowned Sparrow",  "White-throated Sparrow", "Winter Wren", "Yellow-bellied Sapsucker", "Yellow-rumped Warbler"
)
# "Eastern Bluebird",  "Pacific-slope Flycatcher","White-headed Woodpecker",
missing_names <- positive_short[!(positive_short %in% traits$common_name)] #do all these names appear in the BC list?

negative_short <- c("Alder Flycatcher", "Anna's Hummingbird", "Barred Owl", "Black-headed Grosbeak", "Black-throated Gray Warbler", "Boreal Chickadee", "Brown Creeper", "Cassin's Vireo", "Chestnut-backed Chickadee", "Common Yellowthroat", "Dusky Flycatcher", "Golden-crowned Kinglet", "Great Horned Owl", "Hammond's Flycatcher", "Hermit Thrush", "MacGillivray's Warbler", "Mountain Chickadee",  "Nashville Warbler", "Northern Flicker", "Northern Saw-whet Owl", "Northern Waterthrush", "Orange-crowned Warbler", "Ovenbird", "Pygmy Nuthatch", "Red-breasted Nuthatch", "Red-eyed Vireo", "Red-naped Sapsucker", "Ruby-crowned Kinglet", "Ruffed Grouse", "Spotted Towhee", "Swainson's Thrush", "Swamp Sparrow", "Townsend's Warbler", "Warbling Vireo", "Western Wood-Pewee", "Yellow-rumped Warbler"
)
#"Acorn Woodpecker", "Mountain Quail",
missing_names <- negative_short[!(negative_short %in% traits$common_name)] #do all these names appear in the BC list?

null_short <- c("American Goldfinch", "American Robin", "American Wigeon", "Black-capped Chickadee", "Black-headed Grosbeak", "Black-throated Gray Warbler", "Brown Creeper", "Brown-headed Cowbird", "Bufflehead", "Calliope Hummingbird", "Canvasback", "Cassin's Finch", "Cassin's Vireo", "Cedar Waxwing", "Chipping Sparrow", "Clark's Nutcracker", "Common Merganser", "Common Raven", "Common Yellowthroat", "Dark-eyed Junco", "Downy Woodpecker", "Dusky Flycatcher", "Eastern Phoebe", "Evening Grosbeak", "Fox Sparrow", "Golden-crowned Kinglet", "Green-winged Teal", "Hammond's Flycatcher", "Hermit Thrush", "House Finch", "Lesser Scaup", "Lincoln's Sparrow", "MacGillivray's Warbler", "Mallard", "Marsh Wren", "Mountain Chickadee", "Mourning Dove", "Nashville Warbler", "Northern Pintail", "Northern Shoveler", "Northern Waterthrush", "Olive-sided Flycatcher", "Orange-crowned Warbler", "Pacific Wren", "Pine Grosbeak", "Pine Siskin", "Pileated Woodpecker", "Purple Finch", "Red Crossbill", "Red-breasted Merganser", "Red-breasted Nuthatch", "Red-naped Sapsucker", "Red-tailed Hawk", "Ring-necked Duck", "Rock Wren", "Ruby-crowned Kinglet", "Ruffed Grouse", "Rufous Hummingbird", "Song Sparrow", "Spotted Towhee", "Steller's Jay", "Surf Scoter", "Swainson's Thrush", "Townsend's Warbler", "Varied Thrush", "Violet-green Swallow", "Warbling Vireo", "Western Tanager", "Western Wood-Pewee", "White-breasted Nuthatch", "White-crowned Sparrow", "White-winged Scoter", "Williamson's Sapsucker", "Willow Flycatcher", "Wilson's Warbler", "Winter Wren", "Yellow Warbler", "Yellow-bellied Flycatcher", "Yellow-rumped Warbler"
)
# "Greater Scaup","Spotted Sandpiper","Acorn Woodpecker","Mountain Quail","Hermit Warbler",
missing_names <- null_short[!(null_short %in% traits$common_name)] #do all these names appear in the BC list?

traits$positive_short <- ifelse(traits$common_name %in% positive_short, 1, NA)
traits$negative_short <- ifelse(traits$common_name %in% negative_short, 1, NA)
traits$null_short <- ifelse(traits$common_name %in% null_short, 1, NA)


traits <- traits %>%
    filter(!if_all(c(positive_short, negative_short, null_short), is.na)) %>% 
    mutate(across(c(positive_short, negative_short, null_short), ~replace_na(.x, 0)))
```

```{r}
trait.pca <- traits %>% 
  select(common_name, where(is.numeric), -elevational_range,-positive_short, -negative_short, -null_short,  -kipps_distance, -repro_age, -clutches_per_year, -lifespan, -eco_spec_index, -ends_with("_id"), -fstimb)
# trait_list <- trait.pca 
# trait_list <- trait_list[1, ]
# write.csv(trait_list, here("Data", "cleaned", "traits.for.pca.csv"), row.names = FALSE)


pca_result <- prcomp(trait.pca[,-1], scale = T) #scale the variables!
##
pca_result$rotation
pca_result$x

biplot(pca_result)

pca_loadings <- as.data.frame(pca_result$rotation) %>% select(1:10)

summary(pca_result)
VE <- pca_result$sdev^2
PVE <- VE / sum(VE)
PVE <- round(PVE, 2)

pve_df <- data.frame(
  PC = 1:length(PVE),
  PVE = PVE
)

(scree <- ggplot(pve_df, aes(x = PC, y = PVE)) +
  geom_col() +
  xlab("Principal Component") +
  ylab("Proportion of Variance Explained") +
  ggtitle("Scree Plot") +
  ylim(0, 1) +
    theme_bw()+
  scale_x_continuous(breaks = 1:length(PVE)))
ggsave(here("Figures", "PCA", "scree plot.short.png"), plot = scree, width = 8, height =8)


##
pca_df <- as.data.frame(pca_result$x) %>%
  bind_cols(traits %>% select(common_name, positive_short, negative_short, null_short))
pca_df <- pca_df %>%
  mutate(fire_response = case_when(
    positive_short == 1 & negative_short == 0 & null_short == 0 ~ "Positive",
    positive_short == 0 & negative_short == 1 & null_short == 0 ~ "Negative",
    positive_short == 0 & negative_short == 0 & null_short == 1 ~ "Neutral",
    positive_short == 1 & negative_short == 1 & null_short == 0 ~ "Positive & Negative",
    positive_short == 1 & null_short == 1 & negative_short == 0 ~ "Positive & Neutral",
    negative_short == 1 & null_short == 1 & positive_short == 0 ~ "Negative & Neutral",
    TRUE ~ "None"
  ))

loadings <- as.data.frame(pca_result$rotation)  # columns = PCs, rows = traits
loadings$trait <- rownames(loadings)

# Optional: scale arrows for visualization
arrow_multiplier <- 5  # adjust to make arrows visible
loadings <- loadings %>%
  mutate(PC1_arrow = PC1 * arrow_multiplier,
         PC2_arrow = PC2 * arrow_multiplier)


(f1 <- ggplot(pca_df, aes(PC1, PC2, color = fire_response)) +
  geom_point(size = 2, alpha = 0.9) +
  theme_classic() +
  labs(
    title = "PCA of Bird Traits Colored by Fire Response",
    x = paste0("PC1 (", round(summary(pca_result)$importance[2,1]*100, 1), "%)"),
    y = paste0("PC2 (", round(summary(pca_result)$importance[2,2]*100, 1), "%)")
  ) +
   geom_segment(data = loadings,
               aes(x = 0, y = 0, xend = PC1_arrow, yend = PC2_arrow),
               arrow = arrow(length = unit(0.3, "cm")),
               color = "black") +
  geom_text(data = loadings,
            aes(x = PC1_arrow * 1.1, y = PC2_arrow * 1.1, label = trait),
            color = "black", size = 3, hjust = 0.5, vjust = 0.5) +
  scale_color_manual(values = c(
    "Positive" = "#005AB5",
    "Negative" = "#DC3220",
    "Neutral" = "grey",
    "Positive & Negative" = "#44AA99",
    "Positive & Neutral" = "#66a61e",
    "Negative & Neutral" = "#e6ab02"
  )))
(f1.1 <- ggplot(pca_df, aes(PC1, PC2, color = fire_response)) +
  geom_point(size = 2, alpha = 0.9) +
  theme_classic() +
  labs(
    title = "PCA of Bird Traits Colored by Fire Response",
    x = paste0("PC1 (", round(summary(pca_result)$importance[2,1]*100, 1), "%)"),
    y = paste0("PC2 (", round(summary(pca_result)$importance[2,2]*100, 1), "%)")
  ) +
   geom_segment(data = loadings,
               aes(x = 0, y = 0, xend = PC1_arrow, yend = PC2_arrow),
               arrow = arrow(length = unit(0.3, "cm")),
               color = "black") +
  geom_text(data = loadings,
            aes(x = PC1_arrow * 1.1, y = PC2_arrow * 1.1, label = trait),
            color = "black", size = 3, hjust = 0.5, vjust = 0.5) +
  scale_color_manual(values = c(
    "Positive" = "#005AB5",
    "Negative" = "#DC3220",
    "Neutral" = "grey",
    "Positive & Negative" = "#44AA99",
    "Positive & Neutral" = "#005AB5",
    "Negative & Neutral" = "#DC3220"
  )))

# ggplot(pca_df, aes(x = PC1, y = PC2, group = common_name)) +
#       geom_point() + # Add points for individual data points
#       stat_ellipse(level = 0.95, type = "t") + # Add 95% confidence ellipses based on a t-distribution
#       labs(title = "PCA Plot with Confidence Ellipses",
#            x = "Principal Component 1",
#            y = "Principal Component 2") +
#       theme_minimal()
```

```{r}
arrow_multiplier <- 3
loadings <- loadings %>%
  mutate(PC1_arrow = PC1 * arrow_multiplier,
         PC2_arrow = PC2 * arrow_multiplier)

# Plot
(f2 <- ggplot() +
  geom_segment(data = loadings,
               aes(x = 0, y = 0, xend = PC1_arrow, yend = PC2_arrow),
               arrow = arrow(length = unit(0.3, "cm")),
               color = "black") +
  geom_text(data = loadings,
            aes(x = PC1_arrow * 1.1, y = PC2_arrow * 1.1, label = trait),
            color = "black", size = 3) +
  theme_bw() +
  labs(x = "PC1", y = "PC2"))
ggsave(here("Figures", "PCA", "pca loadings.short.png"), plot = f2, width = 10, height =8)
```

```{r}
pca_scores <- as.data.frame(pca_result$x)
pca_scores$common_name <- traits$common_name
pca_scores$positive <- traits$positive
pca_scores$negative <- traits$negative
pca_scores$null <- traits$null

# 3️⃣ ggplot biplot colored by positive
(f3 <- ggplot(pca_scores, aes(x = PC1, y = PC2, color = factor(positive))) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_manual(values = c("0" = "grey", "1" = "#005AB5")) +
  theme_bw() +
  labs(color = "Positive Response", x = "PC1", y = "PC2",
       title = "PCA of Bird Traits Colored by Positive Fire Response"))
(f4 <- ggplot(pca_scores, aes(x = PC1, y = PC2, color = factor(negative))) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_manual(values = c("0" = "grey", "1" = "#DC3220")) +
  theme_bw() +
  labs(color = "Negative Response", x = "PC1", y = "PC2",
       title = "PCA of Bird Traits Colored by Negative Fire Response"))

(pos.neg <- f3/f4)
ggsave(here("Figures", "PCA", "pca by response.png"), plot = pos.neg, width = 8, height =10)


##
(f5 <- ggplot(pca_scores, aes(x = PC1, y = PC2, color = factor(positive))) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_manual(values = c("0" = "grey", "1" = "#005AB5")) +
  theme_bw() +
  geom_text(aes(label = common_name), size = 3) +
  labs(color = "Positive Response", x = "PC1", y = "PC2",
       title = "PCA of Bird Traits Colored by Positive Fire Response"))
(f6 <- ggplot(pca_scores, aes(x = PC1, y = PC2, color = factor(negative))) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_manual(values = c("0" = "grey", "1" = "#DC3220")) +
    geom_text(aes(label = common_name), size = 3) +
  theme_bw() +
  labs(color = "Negative Response", x = "PC1", y = "PC2",
       title = "PCA of Bird Traits Colored by Negative Fire Response"))

(pos.neg.spp.names <- f5/f6)
ggsave(here("Figures", "PCA", "pca by response spp names.short.png"), plot = pos.neg.spp.names, width = 20, height =20, dpi = 300)

```

```{r}
# Extract loadings
loadings <- as.data.frame(pca_result$rotation)

# Add trait names as a column
loadings$trait <- rownames(loadings)

loadings_long <- loadings %>%
  pivot_longer(cols = starts_with("PC"), 
               names_to = "PC", 
               values_to = "loading") %>%
  filter(PC %in% paste0("PC", 1:2))   # keep only PCs 1-5

(loadings.bar <- ggplot(loadings_long, aes(x = trait, y = loading, fill = PC)) +
  geom_col(position = "dodge") +   # side-by-side bars per PC
  coord_flip() +                   # horizontal bars
  labs(x = "Trait", y = "Loading") +
  theme_bw(base_size = 14) +
  scale_y_continuous(limits = c(-0.5, 0.5), breaks = seq(-0.5, 0.5, 0.25)) +  # set range and breaks
  scale_fill_manual(values = c("#a1d99b", "#688c64")) +  # set all bars to green
  facet_wrap(~PC)+
    theme(legend.position = "none"))

ggsave(here("Figures", "PCA", "loadings bar plot.short.png"), plot = loadings.bar, width = 6, height =5, dpi = 300)
loadings %>%  arrange(desc(abs(PC1)))

```



```{r}
bc <- traits2 %>%  
  select(common_name, where(is.numeric), -elevational_range,  -kipps_distance, -repro_age, -clutches_per_year, -lifespan, -eco_spec_index, -ends_with("_id"), -fstimb) %>% 
  # mutate(null = NA,
  #        positive = NA,
  #        negative = NA) %>% 
  drop_na()  

#plotting all spp on this pca:
bc_scaled <- scale(bc[,-1],
                    center = (pca_result$center),
                    scale = (pca_result$scale))

bc_scores <- as.data.frame(as.matrix(bc_scaled) %*% pca_result$rotation)

```

```{r}

bc_scores$common_name <- bc$common_name
pca_scores <- pca_scores %>%
  mutate(fire_response = case_when(
    positive == 1 & negative == 1 ~ "mixed",
    positive == 1 & negative == 0 ~ "positive",
    negative == 1 & positive == 0 ~ "negative",
    TRUE ~ "neutral"
  ))
centers <- data.frame(
  group = c("positive", "mixed", "negative", "neutral"),
  PC1 = c(
    mean(pca_scores$PC1[pca_scores$positive == 1 & pca_scores$negative != 1], na.rm = TRUE),
    mean(pca_scores$PC1[pca_scores$positive == 1 & pca_scores$negative == 1], na.rm = TRUE),
    mean(pca_scores$PC1[pca_scores$negative == 1 & pca_scores$positive != 1], na.rm = TRUE),
    mean(pca_scores$PC1[pca_scores$null == 1 & pca_scores$positive == 0 & pca_scores$negative == 0], na.rm = TRUE)),
  PC2 = c(
    mean(pca_scores$PC2[pca_scores$positive == 1 & pca_scores$negative != 1], na.rm = TRUE),
    mean(pca_scores$PC2[pca_scores$positive == 1 & pca_scores$negative == 1], na.rm = TRUE),
    mean(pca_scores$PC2[pca_scores$negative == 1 & pca_scores$positive != 1], na.rm = TRUE),
    mean(pca_scores$PC2[pca_scores$null == 1 & pca_scores$positive == 0 & pca_scores$negative == 0], na.rm = TRUE)
  )
)
# Basic PCA scatterplot
(all <- ggplot(bc_scores, aes(x = PC1, y = PC2)) +  # V1 and V2 are PC1 and PC2
  geom_point(size = 3, alpha = 0.7) +
  labs(x = "PC1", y = "PC2", title = "All BC Birds Plotted on PCA Axes") +
  theme_bw())
ggsave(here("Figures", "PCA", "all spp.short.png"), plot = all, width = 15, height =15, dpi = 300)


# Original PCA points
orig_df <- pca_scores
orig_df$type <- "PCA"


# New projected points
new_df <- bc_scores
new_df$common_name <- bc$common_name[complete.cases(bc)]

# Combine
plot_df <- left_join(new_df,orig_df)

# Overlay plot
(p1 <- ggplot(orig_df, aes(x = PC1, y = PC2)) +
    geom_point(data = new_df, aes( x = PC1, y = PC2), shape = 1, color = "grey", alpha = 0.85, size = 2)+
  geom_point(aes(color = factor(positive)), size = 2.5, alpha = 0.8) +
  scale_color_manual(values = c("0" = "grey50", "1" = "#005AB5")) +
  theme_bw() +
  labs(x = "PC1", y = "PC2", color = "Positive Response", shape = "Point Type"))

(n1 <- ggplot(orig_df, aes(x = PC1, y = PC2)) +
    geom_point(data = new_df, aes( x = PC1, y = PC2), shape = 1, color = "grey", alpha = 0.85, size = 2)+
  geom_point(aes(color = factor(negative)), size = 2.5, alpha = 0.8) +
  scale_color_manual(values = c("0" = "grey50", "1" = "#D55E00")) +
  theme_bw() +
  labs(x = "PC1", y = "PC2", color = "Negative Response", shape = "Point Type"))

orig_df <- orig_df %>% 
  mutate(
    mixed = rowSums(data.frame(positive, negative), na.rm = TRUE),
    mixed = ifelse(mixed == 2|mixed == 3, 1, 0))

(m1 <- ggplot(orig_df, aes(x = PC1, y = PC2)) +
    geom_point(data = new_df, aes( x = PC1, y = PC2), shape = 1, color = "grey", alpha = 0.85, size = 2)+
  geom_point(aes(color = factor(mixed)), size = 2.5, alpha = 0.8) +
  scale_color_manual(values = c("0" = "grey50", "1" = "darkmagenta")) +
  theme_bw() +
  labs(x = "PC1", y = "PC2", color = "Mixed Response", shape = "Point Type"))

(pcas <- p1/n1/m1 + plot_annotation(tag_levels = 'A')
)
ggsave(here("Figures", "PCA", "PCA_all response patchwork.short.png"), plot = pcas, width = 8, height =12, dpi = 300)


(with.mix <- ggplot(pca_scores) +
  geom_point(data = new_df, aes( x = PC1, y = PC2), shape = 1, color = "grey", alpha = 0.85, size = 2)+
  geom_point(aes(PC1, PC2), color = "grey50") +
  geom_point(data = subset(pca_scores, positive == 1),
             aes(PC1, PC2), color = "#0072B2", size = 3, alpha = 0.8) +
  geom_point(data = subset(pca_scores, negative == 1),
             aes(PC1, PC2), color = "#D55E00", size = 3, alpha = 0.8) +
  geom_point(data = subset(pca_scores, negative == 1& positive == 1),
             aes(PC1, PC2), color = "darkmagenta", size = 3, alpha = 0.8)+
  geom_point(data = centers, aes(PC1, PC2), size = 5, color = "white", shape = 16) +
  geom_point(data = centers, aes(PC1, PC2, color = group), size = 7, shape = 8) +
  scale_color_manual(values = c("darkmagenta", "#D55E00","grey50","#0072B2")) +
  theme_bw() +
  theme(legend.position="bottom")+
  labs(color = "Fire Response",
       title = ""))
ggsave(here("Figures", "PCA", "PCA_centers.short.png"), plot = with.mix, width = 8, height =8, dpi = 300)

# t.test(
#   pca_scores$PC1[pca_scores$positive == 1&pca_scores$negative != 1],
#   pca_scores$PC1[pca_scores$negative == 1&pca_scores$positive != 1]
# )
# 
# # PC2
# t.test(
#   pca_scores$PC2[pca_scores$positive == 1&pca_scores$negative != 1],
#   pca_scores$PC2[pca_scores$negative == 1&pca_scores$positive != 1]
# )

anova_pc1 <- aov(PC1 ~ fire_response, data = pca_scores)
summary(anova_pc1)

# Post-hoc (Tukey) test for pairwise comparisons
TukeyHSD(anova_pc1)

anova_pc2 <- aov(PC2~ fire_response, data = pca_scores)
summary(anova_pc2)

# Post-hoc (Tukey) test for pairwise comparisons
TukeyHSD(anova_pc2)

#p < 0.05 → The positive and negative 1’s occupy significantly different mean positions along that PC.
#p ≥ 0.05 → No significant difference.
arrow_multiplier <- 5  # adjust to make arrows visible
loadings <- loadings %>%
  mutate(PC1_arrow = PC1 * arrow_multiplier,
         PC2_arrow = PC2 * arrow_multiplier)
(f2 <- ggplot() +
  geom_segment(data = loadings,
               aes(x = 0, y = 0, xend = PC1_arrow, yend = PC2_arrow),
               arrow = arrow(length = unit(0.3, "cm")),
               color = "black") +
  geom_text_repel(data = loadings,
                  aes(x = PC1_arrow, y = PC2_arrow, label = trait),
                  color = "black",
                  size = 4,
                  segment.color = "grey50",    # line from label to point
                  segment.size = 0.5,
                  segment.alpha = 0.8,
                  nudge_x = 0,                 # optional small nudge
                  nudge_y = 0,
                  box.padding = 0.5,
                  point.padding = 0.5,
                  max.overlaps = 20) +
  geom_point(data = centers, aes(PC1, PC2, color = group), size = 6, shape = 8) +
  theme_bw() +
  theme(legend.position = "none")+
  scale_color_manual(values = c("darkmagenta", "#D55E00","grey50","#0072B2")) +
  labs(x = paste0("PC1 (", round(summary(pca_result)$importance[2,1]*100, 1), "%)"),
    y = paste0("PC2 (", round(summary(pca_result)$importance[2,2]*100, 1), "%)")))

(short <- with.mix+f2+  
  plot_layout(guides = "collect") +
  plot_annotation(theme = theme(legend.position = "bottom"), tag_levels = "A"))
ggsave(here("Figures", "PCA", "PCA_end fig short.png"), plot = short, width = 18, height =8, dpi = 300)

```
> anova_pc1 <- aov(PC1 ~ fire_response, data = pca_scores)
> summary(anova_pc1)
               Df Sum Sq Mean Sq F value   Pr(>F)    
fire_response   3     88  29.346   5.947 0.000807 ***
Residuals     122    602   4.934                     
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
> # Post-hoc (Tukey) test for pairwise comparisons
> TukeyHSD(anova_pc1)
  Tukey multiple comparisons of means
    95% family-wise confidence level

Fit: aov(formula = PC1 ~ fire_response, data = pca_scores)

$fire_response
                        diff        lwr        upr     p adj
negative-mixed    -0.7273669 -2.8038070  1.3490733 0.7982501
neutral-mixed     -2.6291953 -4.5451468 -0.7132438 0.0028012
positive-mixed    -1.5079571 -3.3545297  0.3386156 0.1503312
neutral-negative  -1.9018284 -3.4519095 -0.3517473 0.0094896
positive-negative -0.7805902 -2.2440490  0.6828686 0.5084192
positive-neutral   1.1212382 -0.1039059  2.3463823 0.0856007

> anova_pc2 <- aov(PC2~ fire_response, data = pca_scores)
> summary(anova_pc2)
               Df Sum Sq Mean Sq F value  Pr(>F)   
fire_response   3   40.9  13.617   4.443 0.00533 **
Residuals     122  373.9   3.065                   
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
> 
> # Post-hoc (Tukey) test for pairwise comparisons
> TukeyHSD(anova_pc2)
  Tukey multiple comparisons of means
    95% family-wise confidence level

Fit: aov(formula = PC2 ~ fire_response, data = pca_scores)

$fire_response
                        diff        lwr        upr     p adj
negative-mixed    -0.6174821 -2.2539641  1.0190000 0.7594761
neutral-mixed     -1.7025504 -3.2125482 -0.1925525 0.0204195
positive-mixed    -1.5777405 -3.0330595 -0.1224214 0.0280118
neutral-negative  -1.0850683 -2.3067167  0.1365801 0.1005472
positive-negative -0.9602584 -2.1136382  0.1931213 0.1378861
positive-neutral   0.1248099 -0.8407495  1.0903692 0.9867910
