---
title: "pca_medium"
author: "JL"
date: "`r Sys.Date()`"
output: html_document
---

```{r}
library(here)
library(tidyverse)
library(patchwork)
```

```{r medium TERM}
traits <- read.csv(here("Data", "cleaned", "BC_bird_traits.csv"))
traits$habitat_density[traits$habitat_density == "dense"] <- 1
traits$habitat_density[traits$habitat_density == "semiopen"] <- 2
traits$habitat_density[traits$habitat_density == "open"] <- 3
traits <- traits %>% 
   mutate(cavity_nester = ifelse(nest_category == "cavity", 1, 0),
          ground_nester = ifelse(most_common_nesting_location == "ground", 1, 0),
          shrub_nester = ifelse(most_common_nesting_location == "shrub", 1, 0)) %>%
    mutate(habitat_density = as.numeric(habitat_density)) 


##
positive_medium <- c("Alder Flycatcher", "Acorn Woodpecker", "American Kestrel", "American Robin",  "Black-capped Chickadee", "Black-throated Gray Warbler", "Brewer's Blackbird", "Brewer's Sparrow", "Brown Creeper", "Blue-headed Vireo", "Canada Jay", "Canada Goose", "Cassin's Finch", "Cedar Waxwing", "Chestnut-sided Warbler", "Chipping Sparrow", "Clark's Nutcracker", "Connecticut Warbler", "Common Yellowthroat", "Dark-eyed Junco", "Downy Woodpecker", "Dusky Flycatcher", "Eastern Bluebird", "Fox Sparrow", "Green-tailed Towhee", "Hairy Woodpecker", "Hermit Thrush", "Hermit Warbler", "Northern House Wren", "Lazuli Bunting", "Least Flycatcher", "Lewis's Woodpecker", "Lincoln's Sparrow", "MacGillivray's Warbler", "Mountain Chickadee", "Mountain Quail", "Nashville Warbler", "Northern Flicker", "Northern Pygmy-Owl", "Pine Siskin", "Redpoll", "Red-breasted Nuthatch", "Red-winged Blackbird", "Rock Wren", "Rufous Hummingbird", "Savannah Sparrow", "Say's Phoebe", "Song Sparrow", "Spotted Towhee", "Steller's Jay", "Swamp Sparrow", "Tennessee Warbler", "Townsend's Solitaire","Tree Swallow","Turkey Vulture", "Veery", "Warbling Vireo", "Western Screech-Owl", "Western Tanager", "Western Wood-Pewee", "White-breasted Nuthatch", "White-crowned Sparrow", "White-headed Woodpecker", "White-throated Sparrow", "Winter Wren", "Yellow-bellied Sapsucker", "Yellow-rumped Warbler")
missing_names <- positive_medium[!(positive_medium %in% traits$common_name)] #do all these names appear in the BC list?

negative_medium <- c("Alder Flycatcher", "Barred Owl", "Black-headed Grosbeak", "Black-throated Gray Warbler", "Blue-headed Vireo", "Boreal Chickadee", "Brown Creeper", "Cassin's Vireo", "Chestnut-backed Chickadee", "Common Yellowthroat", "Dusky Flycatcher", "Golden-crowned Kinglet", "Great Horned Owl", "Hermit Thrush", "Hermit Warbler", "MacGillivray's Warbler", "Mountain Chickadee", "Nashville Warbler", "Northern Saw-whet Owl", "Northern Waterthrush", "Orange-crowned Warbler", "Ovenbird", "Pygmy Nuthatch", "Red-breasted Nuthatch", "Red-eyed Vireo", "Red-naped Sapsucker", "Ruby-crowned Kinglet", "Ruffed Grouse", "Ruby-crowned Kinglet", "Swainson's Thrush", "Swamp Sparrow", "Townsend's Warbler", "Warbling Vireo", "Western Wood-Pewee", "Yellow-rumped Warbler")
missing_names <- negative_medium[!(negative_medium %in% traits$common_name)] #do all these names appear in the BC list?

null_medium <- c("American Goldfinch", "American Robin", "American Black Duck", "American Wigeon", "Black-capped Chickadee", "Black-headed Grosbeak", "Bufflehead", "Brown Creeper", "Brown-headed Cowbird", "Calliope Hummingbird", "Canvasback", "Cassin's Finch", "Cedar Waxwing", "Chipping Sparrow", "Clark's Nutcracker", "Common Merganser", "Common Goldeneye", "Common Loon", "Common Raven", "Common Yellowthroat", "Dark-eyed Junco", "Downy Woodpecker", "Dusky Flycatcher", "Eastern Phoebe", "Evening Grosbeak", "Fox Sparrow",  "Greater Scaup", "Green-winged Teal", "Hammond's Flycatcher", "Hermit Thrush", "House Finch", "Lesser Scaup", "Lincoln's Sparrow", "MacGillivray's Warbler", "Mallard", "Marsh Wren", "Mountain Chickadee", "Mourning Dove", "Nashville Warbler", "Northern Pintail", "Northern Shoveler", "Northern Waterthrush", "Olive-sided Flycatcher", "Orange-crowned Warbler", "Pacific Wren", "Pine Grosbeak", "Pine Siskin", "Pileated Woodpecker", "Red Crossbill", "Red-breasted Merganser", "Red-breasted Nuthatch", "Red-naped Sapsucker", "Red-tailed Hawk", "Ring-necked Duck", "Rock Wren", "Ruby-crowned Kinglet", "Ruffed Grouse", "Rufous Hummingbird", "Song Sparrow", "Spotted Sandpiper", "Steller's Jay", "Surf Scoter", "Swainson's Thrush", "Townsend's Warbler", "Varied Thrush", "Violet-green Swallow", "Warbling Vireo", "Western Tanager", "White-breasted Nuthatch", "White-crowned Sparrow", "White-winged Scoter", "Willow Flycatcher", "Wilson's Warbler", "Williamson's Sapsucker", "Winter Wren", "Yellow Warbler", "Yellow-bellied Flycatcher", "Yellow-rumped Warbler")
missing_names <- null_medium[!(null_medium %in% traits$common_name)] #do all these names appear in the BC list?

traits$positive_medium <- ifelse(traits$common_name %in% positive_medium, 1, NA)
traits$negative_medium <- ifelse(traits$common_name %in% negative_medium, 1, NA)
traits$null_medium <- ifelse(traits$common_name %in% null_medium, 1, NA)


traits <- traits %>%
    filter(!if_all(c(positive_medium, negative_medium, null_medium), is.na)) %>% 
    mutate(across(c(positive_medium, negative_medium, null_medium), ~replace_na(.x, 0)))
```

```{r}
trait.pca <- traits %>% select(common_name, where(is.numeric), -elevational_range,-positive_medium, -negative_medium, -null_medium, -nest_tree_use, -kipps_distance, -repro_age, -clutches_per_year)

# trait_list <- trait.pca 
# trait_list <- trait_list[1, ]
# write.csv(trait_list, here("Data", "cleaned", "traits.for.pca.csv"), row.names = FALSE)


pca_result <- prcomp(trait.pca[,-1], scale = T) #scale the variables!
##
pca_result$rotation
pca_result$x

biplot(pca_result)

pca_loadings <- as.data.frame(pca_result$rotation) %>% select(1:10)

summary(pca_result)
VE <- pca_result$sdev^2
PVE <- VE / sum(VE)
PVE <- round(PVE, 2)

pve_df <- data.frame(
  PC = 1:length(PVE),
  PVE = PVE
)

(scree <- ggplot(pve_df, aes(x = PC, y = PVE)) +
  geom_col() +
  xlab("Principal Component") +
  ylab("Proportion of Variance Explained") +
  ggtitle("Scree Plot") +
  ylim(0, 1) +
    theme_bw()+
  scale_x_continuous(breaks = 1:length(PVE)))
ggsave(here("Figures", "PCA", "scree plot.medium.png"), plot = scree, width = 8, height =8)


##
pca_df <- as.data.frame(pca_result$x) %>%
  bind_cols(traits %>% select(common_name, positive_medium, negative_medium, null_medium))
pca_df <- pca_df %>%
  mutate(fire_response = case_when(
    positive_medium == 1 & negative_medium == 0 & null_medium == 0 ~ "Positive",
    positive_medium == 0 & negative_medium == 1 & null_medium == 0 ~ "Negative",
    positive_medium == 0 & negative_medium == 0 & null_medium == 1 ~ "Neutral",
    positive_medium == 1 & negative_medium == 1 & null_medium == 0 ~ "Positive & Negative",
    positive_medium == 1 & null_medium == 1 & negative_medium == 0 ~ "Positive & Neutral",
    negative_medium == 1 & null_medium == 1 & positive_medium == 0 ~ "Negative & Neutral",
    TRUE ~ "None"
  ))

loadings <- as.data.frame(pca_result$rotation)  # columns = PCs, rows = traits
loadings$trait <- rownames(loadings)

# Optional: scale arrows for visualization
arrow_multiplier <- 5  # adjust to make arrows visible
loadings <- loadings %>%
  mutate(PC1_arrow = PC1 * arrow_multiplier,
         PC2_arrow = PC2 * arrow_multiplier)


(f1 <- ggplot(pca_df, aes(PC1, PC2, color = fire_response)) +
  geom_point(size = 2, alpha = 0.9) +
  theme_classic() +
  labs(
    title = "PCA of Bird Traits Colored by Fire Response",
    x = paste0("PC1 (", round(summary(pca_result)$importance[2,1]*100, 1), "%)"),
    y = paste0("PC2 (", round(summary(pca_result)$importance[2,2]*100, 1), "%)")
  ) +
   geom_segment(data = loadings,
               aes(x = 0, y = 0, xend = PC1_arrow, yend = PC2_arrow),
               arrow = arrow(length = unit(0.3, "cm")),
               color = "black") +
  geom_text(data = loadings,
            aes(x = PC1_arrow * 1.1, y = PC2_arrow * 1.1, label = trait),
            color = "black", size = 3, hjust = 0.5, vjust = 0.5) +
  scale_color_manual(values = c(
    "Positive" = "#005AB5",
    "Negative" = "#DC3220",
    "Neutral" = "grey",
    "Positive & Negative" = "#44AA99",
    "Positive & Neutral" = "#66a61e",
    "Negative & Neutral" = "#e6ab02"
  )))
(f1.1 <- ggplot(pca_df, aes(PC1, PC2, color = fire_response)) +
  geom_point(size = 2, alpha = 0.9) +
  theme_classic() +
  labs(
    title = "PCA of Bird Traits Colored by Fire Response",
    x = paste0("PC1 (", round(summary(pca_result)$importance[2,1]*100, 1), "%)"),
    y = paste0("PC2 (", round(summary(pca_result)$importance[2,2]*100, 1), "%)")
  ) +
   geom_segment(data = loadings,
               aes(x = 0, y = 0, xend = PC1_arrow, yend = PC2_arrow),
               arrow = arrow(length = unit(0.3, "cm")),
               color = "black") +
  geom_text(data = loadings,
            aes(x = PC1_arrow * 1.1, y = PC2_arrow * 1.1, label = trait),
            color = "black", size = 3, hjust = 0.5, vjust = 0.5) +
  scale_color_manual(values = c(
    "Positive" = "#005AB5",
    "Negative" = "#DC3220",
    "Neutral" = "grey",
    "Positive & Negative" = "#44AA99",
    "Positive & Neutral" = "#005AB5",
    "Negative & Neutral" = "#DC3220"
  )))

# ggplot(pca_df, aes(x = PC1, y = PC2, group = common_name)) +
#       geom_point() + # Add points for individual data points
#       stat_ellipse(level = 0.95, type = "t") + # Add 95% confidence ellipses based on a t-distribution
#       labs(title = "PCA Plot with Confidence Ellipses",
#            x = "Principal Component 1",
#            y = "Principal Component 2") +
#       theme_minimal()
```

```{r}
arrow_multiplier <- 3
loadings <- loadings %>%
  mutate(PC1_arrow = PC1 * arrow_multiplier,
         PC2_arrow = PC2 * arrow_multiplier)

# Plot
(f2 <- ggplot() +
  geom_segment(data = loadings,
               aes(x = 0, y = 0, xend = PC1_arrow, yend = PC2_arrow),
               arrow = arrow(length = unit(0.3, "cm")),
               color = "black") +
  geom_text(data = loadings,
            aes(x = PC1_arrow * 1.1, y = PC2_arrow * 1.1, label = trait),
            color = "black", size = 3) +
  theme_bw() +
  labs(x = "PC1", y = "PC2"))
ggsave(here("Figures", "PCA", "pca loadings.medium.png"), plot = f2, width = 10, height =8)
```

```{r}
pca_scores <- as.data.frame(pca_result$x)
pca_scores$common_name <- traits$common_name
pca_scores$positive <- traits$positive
pca_scores$negative <- traits$negative
pca_scores$null <- traits$null

# 3️⃣ ggplot biplot colored by positive
(f3 <- ggplot(pca_scores, aes(x = PC1, y = PC2, color = factor(positive))) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_manual(values = c("0" = "grey", "1" = "#005AB5")) +
  theme_bw() +
  labs(color = "Positive Response", x = "PC1", y = "PC2",
       title = "PCA of Bird Traits Colored by Positive Fire Response"))
(f4 <- ggplot(pca_scores, aes(x = PC1, y = PC2, color = factor(negative))) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_manual(values = c("0" = "grey", "1" = "#DC3220")) +
  theme_bw() +
  labs(color = "Negative Response", x = "PC1", y = "PC2",
       title = "PCA of Bird Traits Colored by Negative Fire Response"))

(pos.neg <- f3/f4)
ggsave(here("Figures", "PCA", "pca by response.png"), plot = pos.neg, width = 8, height =10)


##
(f5 <- ggplot(pca_scores, aes(x = PC1, y = PC2, color = factor(positive))) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_manual(values = c("0" = "grey", "1" = "#005AB5")) +
  theme_bw() +
  geom_text(aes(label = common_name), size = 3) +
  labs(color = "Positive Response", x = "PC1", y = "PC2",
       title = "PCA of Bird Traits Colored by Positive Fire Response"))
(f6 <- ggplot(pca_scores, aes(x = PC1, y = PC2, color = factor(negative))) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_manual(values = c("0" = "grey", "1" = "#DC3220")) +
    geom_text(aes(label = common_name), size = 3) +
  theme_bw() +
  labs(color = "Negative Response", x = "PC1", y = "PC2",
       title = "PCA of Bird Traits Colored by Negative Fire Response"))

(pos.neg.spp.names <- f5/f6)
ggsave(here("Figures", "PCA", "pca by response spp names.medium.png"), plot = pos.neg.spp.names, width = 20, height =20, dpi = 300)

```

```{r}
# Extract loadings
loadings <- as.data.frame(pca_result$rotation)

# Add trait names as a column
loadings$trait <- rownames(loadings)

loadings_long <- loadings %>%
  pivot_longer(cols = starts_with("PC"), 
               names_to = "PC", 
               values_to = "loading") %>%
  filter(PC %in% paste0("PC", 1:5))   # keep only PCs 1-5

(loadings.bar <- ggplot(loadings_long, aes(x = trait, y = loading, fill = PC)) +
  geom_col(position = "dodge") +   # side-by-side bars per PC
  coord_flip() +                   # horizontal bars
  labs(x = "Trait", y = "Loading", 
       title = "PCA Loadings per Trait") +
  theme_bw(base_size = 14) +
  scale_fill_brewer(palette = "Set2")+facet_wrap(~PC)+
    theme(legend.position = "none"))

ggsave(here("Figures", "PCA", "loadings bar plot.medium.png"), plot = loadings.bar, width = 15, height =15, dpi = 300)

```

```{r}
bc <- traits %>% select(common_name, where(is.numeric), -elevational_range, -nest_tree_use, -kipps_distance, -repro_age, -clutches_per_year, -null_medium, -negative_medium, -positive_medium) %>% 
  # mutate(null = NA,
  #        positive = NA,
  #        negative = NA) %>% 
  filter(if_all(1:22, ~ !is.na(.)))

#plotting all spp on this pca:
bc_scaled <- scale(bc[,-1],
                    center = (pca_result$center),
                    scale = (pca_result$scale))

bc_scores <- as.data.frame(as.matrix(bc_scaled) %*% pca_result$rotation)

```

```{r}

bc_scores$common_name <- bc$common_name
centers <- data.frame(
  group = c("positive_short", "negative_short"),
  PC1 = c(
    mean(pca_scores$PC1[pca_scores$positive == 1 & pca_scores$negative != 1], na.rm = TRUE),
    mean(pca_scores$PC1[pca_scores$negative == 1 & pca_scores$positive != 1], na.rm = TRUE)),
  PC2 = c(
    mean(pca_scores$PC2[pca_scores$positive == 1 & pca_scores$negative != 1], na.rm = TRUE),
    mean(pca_scores$PC2[pca_scores$negative == 1 & pca_scores$positive != 1], na.rm = TRUE)
  )
)

# Basic PCA scatterplot
(all <- ggplot(bc_scores, aes(x = PC1, y = PC2)) +  # V1 and V2 are PC1 and PC2
  geom_point(size = 3, alpha = 0.7) +
  labs(x = "PC1", y = "PC2", title = "All BC Birds Plotted on PCA Axes") +
  theme_bw())
ggsave(here("Figures", "PCA", "all spp.medium.png"), plot = all, width = 15, height =15, dpi = 300)


# Original PCA points
orig_df <- pca_scores
orig_df$type <- "PCA"


# New projected points
new_df <- bc_scores
new_df$common_name <- bc$common_name[complete.cases(bc)]

# Combine
plot_df <- left_join(new_df,orig_df)

# Overlay plot
(p1 <- ggplot(orig_df, aes(x = PC1, y = PC2)) +
    geom_point(data = new_df, aes( x = PC1, y = PC2), shape = 1, color = "grey", alpha = 0.85, size = 2)+
  geom_point(aes(color = factor(positive)), size = 2.5, alpha = 0.8) +
  scale_color_manual(values = c("0" = "grey50", "1" = "#005AB5")) +
  theme_bw() +
  labs(x = "PC1", y = "PC2", color = "Positive Response", shape = "Point Type"))

(n1 <- ggplot(orig_df, aes(x = PC1, y = PC2)) +
    geom_point(data = new_df, aes( x = PC1, y = PC2), shape = 1, color = "grey", alpha = 0.85, size = 2)+
  geom_point(aes(color = factor(negative)), size = 2.5, alpha = 0.8) +
  scale_color_manual(values = c("0" = "grey50", "1" = "#D55E00")) +
  theme_bw() +
  labs(x = "PC1", y = "PC2", color = "Negative Response", shape = "Point Type"))

orig_df <- orig_df %>% 
  mutate(
    medium = rowSums(data.frame(positive, negative), na.rm = TRUE),
    medium = ifelse(medium == 2|medium == 3, 1, 0))

(m1 <- ggplot(orig_df, aes(x = PC1, y = PC2)) +
    geom_point(data = new_df, aes( x = PC1, y = PC2), shape = 1, color = "grey", alpha = 0.85, size = 2)+
  geom_point(aes(color = factor(medium)), size = 2.5, alpha = 0.8) +
  scale_color_manual(values = c("0" = "grey50", "1" = "darkmagenta")) +
  theme_bw() +
  labs(x = "PC1", y = "PC2", color = "medium Response", shape = "Point Type"))

(pcas <- p1/n1/m1 + plot_annotation(tag_levels = 'A')
)
ggsave(here("Figures", "PCA", "PCA_all response patchwork.medium.png"), plot = pcas, width = 8, height =12, dpi = 300)


(with.mix <- ggplot(pca_scores) +
  geom_point(data = new_df, aes( x = PC1, y = PC2), shape = 1, color = "grey", alpha = 0.85, size = 2)+
  geom_point(aes(PC1, PC2), color = "grey80") +
  geom_point(data = subset(pca_scores, positive == 1),
             aes(PC1, PC2), color = "#0072B2", size = 3, alpha = 0.8) +
  geom_point(data = subset(pca_scores, negative == 1),
             aes(PC1, PC2), color = "#D55E00", size = 3, alpha = 0.8) +
  geom_point(data = subset(pca_scores, negative == 1& positive == 1),
             aes(PC1, PC2), color = "darkmagenta", size = 3, alpha = 0.8)+
  geom_point(data = centers, aes(PC1, PC2), size = 5, color = "white", shape = 16) +
  geom_point(data = centers, aes(PC1, PC2, color = group), size = 7, shape = 8) +
  scale_color_manual(values = c("#D55E00","#0072B2")) +
  theme_bw() +
  theme(legend.position="bottom")+
  labs(color = "Fire Response",
       title = ""))
ggsave(here("Figures", "PCA", "PCA_centers.medium.png"), plot = with.mix, width = 8, height =8, dpi = 300)

(wo.mix <- ggplot(pca_scores) +
  geom_point(data = new_df, aes( x = PC1, y = PC2), shape = 1, color = "grey", alpha = 0.85, size = 2)+
  geom_point(aes(PC1, PC2), color = "grey80") +
  geom_point(data = subset(pca_scores, positive == 1),
             aes(PC1, PC2), color = "#0072B2", size = 3, alpha = 0.8) +
  geom_point(data = subset(pca_scores, negative == 1),
             aes(PC1, PC2), color = "#D55E00", size = 3, alpha = 0.8) +
  geom_point(data = centers, aes(PC1, PC2), size = 5, color = "white", shape = 16) +
  geom_point(data = centers, aes(PC1, PC2, color = group), size = 5, shape = 8) +
  scale_color_manual(values = c("#D55E00","#0072B2")) +
  theme_bw() +
  labs(color = "Response type",
       title = "Locations of positive vs negative responses in PCA space"))

t.test(
  pca_scores$PC1[pca_scores$positive == 1&pca_scores$negative != 1],
  pca_scores$PC1[pca_scores$negative == 1&pca_scores$positive != 1]
)

# PC2
t.test(
  pca_scores$PC2[pca_scores$positive == 1&pca_scores$negative != 1],
  pca_scores$PC2[pca_scores$negative == 1&pca_scores$positive != 1]
)
#p < 0.05 → The positive and negative 1’s occupy significantly different mean positions along that PC.
#p ≥ 0.05 → No significant difference.
arrow_multiplier <- 5  # adjust to make arrows visible
loadings <- loadings %>%
  mutate(PC1_arrow = PC1 * arrow_multiplier,
         PC2_arrow = PC2 * arrow_multiplier)
(f2 <- ggplot() +
  geom_segment(data = loadings,
               aes(x = 0, y = 0, xend = PC1_arrow, yend = PC2_arrow),
               arrow = arrow(length = unit(0.3, "cm")),
               color = "black") +
  geom_text(data = loadings,
            aes(x = PC1_arrow * 1.1, y = PC2_arrow * 1.1, label = trait),
            color = "black", size = 3) +
  geom_point(data = centers, aes(PC1, PC2, color = group), size = 6, shape = 8) +
  theme_bw() +
  scale_color_manual(values = c("#D55E00","#0072B2")) +
  labs(x = "PC1", y = "PC2"))

(medium <- with.mix+f2
)
ggsave(here("Figures", "PCA", "PCA_end fig medium.png"), plot = medium, width = 12, height =8, dpi = 300)

```



	Welch Two Sample t-test

data:  pca_scores$PC1[pca_scores$positive == 1 & pca_scores$negative != 1] and pca_scores$PC1[pca_scores$negative == 1 & pca_scores$positive != 1]
t = 2.6375, df = 28.282, p-value = 0.01342
alternative hypothesis: true difference in means is not equal to 0
95 percent confidence interval:
 0.3048693 2.4206991
sample estimates:
  mean of x   mean of y 
 0.08768467 -1.27509953 


	Welch Two Sample t-test

data:  pca_scores$PC2[pca_scores$positive == 1 & pca_scores$negative != 1] and pca_scores$PC2[pca_scores$negative == 1 & pca_scores$positive != 1]
t = 0.26589, df = 27.258, p-value = 0.7923
alternative hypothesis: true difference in means is not equal to 0
95 percent confidence interval:
 -0.905831  1.175690
sample estimates:
 mean of x  mean of y 
-0.2647255 -0.3996552