---
title: "pca_medium"
author: "JL"
date: "`r Sys.Date()`"
output: html_document
---

```{r}
library(here)
library(tidyverse)
library(patchwork)
```

```{r medium TERM}
traits <- read.csv(here("Data", "cleaned", "BC_bird_traits.csv"))
traits$habitat_density[traits$habitat_density == "dense"] <- 1
traits$habitat_density[traits$habitat_density == "semiopen"] <- 2
traits$habitat_density[traits$habitat_density == "open"] <- 3
traits <- traits %>% 
    mutate(cavity_nester = ifelse(nest_category == "cavity", 1, 0),
   #        ground_nester = ifelse(most_common_nesting_location == "ground", 1, 0),
   #        shrub_nester = ifelse(most_common_nesting_location == "shrub", 1, 0)) %>%
    habitat_density = as.numeric(habitat_density))
traits2 <- traits

positive_medium <- c("Alder Flycatcher", "Acorn Woodpecker", "American Kestrel", "American Robin",  "Black-capped Chickadee", "Black-throated Gray Warbler", "Brewer's Blackbird", "Brewer's Sparrow", "Brown Creeper", "Blue-headed Vireo", "Canada Jay", "Canada Goose", "Cassin's Finch", "Cedar Waxwing", "Chestnut-sided Warbler", "Chipping Sparrow", "Clark's Nutcracker", "Connecticut Warbler", "Common Yellowthroat", "Dark-eyed Junco", "Downy Woodpecker", "Dusky Flycatcher", "Eastern Bluebird", "Fox Sparrow", "Green-tailed Towhee", "Hairy Woodpecker", "Hermit Thrush", "Hermit Warbler", "Northern House Wren", "Lazuli Bunting", "Least Flycatcher", "Lewis's Woodpecker", "Lincoln's Sparrow", "MacGillivray's Warbler", "Mountain Chickadee", "Mountain Quail", "Nashville Warbler", "Northern Flicker", "Northern Pygmy-Owl", "Pine Siskin", "Redpoll", "Red-breasted Nuthatch", "Red-winged Blackbird", "Rock Wren", "Rufous Hummingbird", "Savannah Sparrow", "Say's Phoebe", "Song Sparrow", "Spotted Towhee", "Steller's Jay", "Swamp Sparrow", "Tennessee Warbler", "Townsend's Solitaire","Tree Swallow","Turkey Vulture", "Veery", "Warbling Vireo", "Western Screech-Owl", "Western Tanager", "Western Wood-Pewee", "White-breasted Nuthatch", "White-crowned Sparrow", "White-headed Woodpecker", "White-throated Sparrow", "Winter Wren", "Yellow-bellied Sapsucker", "Yellow-rumped Warbler")
missing_names <- positive_medium[!(positive_medium %in% traits$common_name)] #do all these names appear in the BC list?

negative_medium <- c("Alder Flycatcher", "Barred Owl", "Black-headed Grosbeak", "Black-throated Gray Warbler", "Blue-headed Vireo", "Boreal Chickadee", "Brown Creeper", "Cassin's Vireo", "Chestnut-backed Chickadee", "Common Yellowthroat", "Dusky Flycatcher", "Golden-crowned Kinglet", "Great Horned Owl", "Hermit Thrush", "Hermit Warbler", "MacGillivray's Warbler", "Mountain Chickadee", "Nashville Warbler", "Northern Saw-whet Owl", "Northern Waterthrush", "Orange-crowned Warbler", "Ovenbird", "Pygmy Nuthatch", "Red-breasted Nuthatch", "Red-eyed Vireo", "Red-naped Sapsucker", "Ruby-crowned Kinglet", "Ruffed Grouse", "Swainson's Thrush", "Swamp Sparrow", "Townsend's Warbler", "Warbling Vireo", "Western Wood-Pewee", "Yellow-rumped Warbler")
missing_names <- negative_medium[!(negative_medium %in% traits$common_name)] #do all these names appear in the BC list?

null_medium <- c("American Goldfinch", "American Robin", "American Black Duck", "American Wigeon", "Black-capped Chickadee", "Black-headed Grosbeak", "Bufflehead", "Brown Creeper", "Brown-headed Cowbird", "Calliope Hummingbird", "Canvasback", "Cassin's Finch", "Cedar Waxwing", "Chipping Sparrow", "Clark's Nutcracker", "Common Merganser", "Common Goldeneye", "Common Loon", "Common Raven", "Common Yellowthroat", "Dark-eyed Junco", "Downy Woodpecker", "Dusky Flycatcher", "Eastern Phoebe", "Evening Grosbeak", "Fox Sparrow",  "Greater Scaup", "Green-winged Teal", "Hammond's Flycatcher", "Hermit Thrush", "House Finch", "Lesser Scaup", "Lincoln's Sparrow", "MacGillivray's Warbler", "Mallard", "Marsh Wren", "Mountain Chickadee", "Mourning Dove", "Nashville Warbler", "Northern Pintail", "Northern Shoveler", "Northern Waterthrush", "Olive-sided Flycatcher", "Orange-crowned Warbler", "Pacific Wren", "Pine Grosbeak", "Pine Siskin", "Pileated Woodpecker", "Red Crossbill", "Red-breasted Merganser", "Red-breasted Nuthatch", "Red-naped Sapsucker", "Red-tailed Hawk", "Ring-necked Duck", "Rock Wren", "Ruby-crowned Kinglet", "Ruffed Grouse", "Rufous Hummingbird", "Song Sparrow", "Spotted Sandpiper", "Steller's Jay", "Surf Scoter", "Swainson's Thrush", "Townsend's Warbler", "Varied Thrush", "Violet-green Swallow", "Warbling Vireo", "Western Tanager", "White-breasted Nuthatch", "White-crowned Sparrow", "White-winged Scoter", "Willow Flycatcher", "Wilson's Warbler", "Williamson's Sapsucker", "Winter Wren", "Yellow Warbler", "Yellow-bellied Flycatcher", "Yellow-rumped Warbler")
missing_names <- null_medium[!(null_medium %in% traits$common_name)] #do all these names appear in the BC list?

traits$positive_medium <- ifelse(traits$common_name %in% positive_medium, 1, NA)
traits$negative_medium <- ifelse(traits$common_name %in% negative_medium, 1, NA)
traits$null_medium <- ifelse(traits$common_name %in% null_medium, 1, NA)


traits <- traits %>%
    filter(!if_all(c(positive_medium, negative_medium, null_medium), is.na)) %>% 
    mutate(across(c(positive_medium, negative_medium, null_medium), ~replace_na(.x, 0)))
```

```{r}
trait.pca <- traits %>% 
  select(common_name, where(is.numeric), -elevational_range,-positive_medium, -negative_medium, -null_medium,  -kipps_distance, -repro_age, -clutches_per_year, -lifespan, -eco_spec_index, -ends_with("_id"), -fstimb)

# trait_list <- trait.pca 
# trait_list <- trait_list[1, ]
# write.csv(trait_list, here("Data", "cleaned", "traits.for.pca.csv"), row.names = FALSE)


pca_result <- prcomp(trait.pca[,-1], scale = T) #scale the variables!
##
pca_result$rotation
pca_result$x

biplot(pca_result)

pca_loadings <- as.data.frame(pca_result$rotation) %>% select(1:10)

summary(pca_result)
VE <- pca_result$sdev^2
PVE <- VE / sum(VE)
PVE <- round(PVE, 2)

pve_df <- data.frame(
  PC = 1:length(PVE),
  PVE = PVE
)

(scree <- ggplot(pve_df, aes(x = PC, y = PVE)) +
  geom_col() +
  xlab("Principal Component") +
  ylab("Proportion of Variance Explained") +
  ggtitle("Scree Plot") +
  ylim(0, 1) +
    theme_bw()+
  scale_x_continuous(breaks = 1:length(PVE)))
ggsave(here("Figures", "PCA", "scree plot.medium.png"), plot = scree, width = 8, height =8)


##
pca_df <- as.data.frame(pca_result$x) %>%
  bind_cols(traits %>% select(common_name, positive_medium, negative_medium, null_medium))
pca_df <- pca_df %>%
  mutate(fire_response = case_when(
    positive_medium == 1 & negative_medium == 0 & null_medium == 0 ~ "Positive",
    positive_medium == 0 & negative_medium == 1 & null_medium == 0 ~ "Negative",
    positive_medium == 0 & negative_medium == 0 & null_medium == 1 ~ "Neutral",
    positive_medium == 1 & negative_medium == 1 & null_medium == 0 ~ "Positive & Negative",
    positive_medium == 1 & null_medium == 1 & negative_medium == 0 ~ "Positive & Neutral",
    negative_medium == 1 & null_medium == 1 & positive_medium == 0 ~ "Negative & Neutral",
    TRUE ~ "None"
  ))

loadings <- as.data.frame(pca_result$rotation)  # columns = PCs, rows = traits
loadings$trait <- rownames(loadings)

# Optional: scale arrows for visualization
arrow_multiplier <- 5  # adjust to make arrows visible
loadings <- loadings %>%
  mutate(PC1_arrow = PC1 * arrow_multiplier,
         PC2_arrow = PC2 * arrow_multiplier)


(f1 <- ggplot(pca_df, aes(PC1, PC2, color = fire_response)) +
  geom_point(size = 2, alpha = 0.9) +
  theme_classic() +
  labs(
    title = "PCA of Bird Traits Colored by Fire Response",
    x = paste0("PC1 (", round(summary(pca_result)$importance[2,1]*100, 1), "%)"),
    y = paste0("PC2 (", round(summary(pca_result)$importance[2,2]*100, 1), "%)")
  ) +
   geom_segment(data = loadings,
               aes(x = 0, y = 0, xend = PC1_arrow, yend = PC2_arrow),
               arrow = arrow(length = unit(0.3, "cm")),
               color = "black") +
  geom_text(data = loadings,
            aes(x = PC1_arrow * 1.1, y = PC2_arrow * 1.1, label = trait),
            color = "black", size = 3, hjust = 0.5, vjust = 0.5) +
  scale_color_manual(values = c(
    "Positive" = "#005AB5",
    "Negative" = "#DC3220",
    "Neutral" = "grey",
    "Positive & Negative" = "#44AA99",
    "Positive & Neutral" = "#66a61e",
    "Negative & Neutral" = "#e6ab02"
  )))
(f1.1 <- ggplot(pca_df, aes(PC1, PC2, color = fire_response)) +
  geom_point(size = 2, alpha = 0.9) +
  theme_classic() +
  labs(
    title = "PCA of Bird Traits Colored by Fire Response",
    x = paste0("PC1 (", round(summary(pca_result)$importance[2,1]*100, 1), "%)"),
    y = paste0("PC2 (", round(summary(pca_result)$importance[2,2]*100, 1), "%)")
  ) +
   geom_segment(data = loadings,
               aes(x = 0, y = 0, xend = PC1_arrow, yend = PC2_arrow),
               arrow = arrow(length = unit(0.3, "cm")),
               color = "black") +
  geom_text(data = loadings,
            aes(x = PC1_arrow * 1.1, y = PC2_arrow * 1.1, label = trait),
            color = "black", size = 3, hjust = 0.5, vjust = 0.5) +
  scale_color_manual(values = c(
    "Positive" = "#005AB5",
    "Negative" = "#DC3220",
    "Neutral" = "grey",
    "Positive & Negative" = "#44AA99",
    "Positive & Neutral" = "#005AB5",
    "Negative & Neutral" = "#DC3220"
  )))

# ggplot(pca_df, aes(x = PC1, y = PC2, group = common_name)) +
#       geom_point() + # Add points for individual data points
#       stat_ellipse(level = 0.95, type = "t") + # Add 95% confidence ellipses based on a t-distribution
#       labs(title = "PCA Plot with Confidence Ellipses",
#            x = "Principal Component 1",
#            y = "Principal Component 2") +
#       theme_minimal()
```

```{r}
arrow_multiplier <- 3
loadings <- loadings %>%
  mutate(PC1_arrow = PC1 * arrow_multiplier,
         PC2_arrow = PC2 * arrow_multiplier)

# Plot
(f2 <- ggplot() +
  geom_segment(data = loadings,
               aes(x = 0, y = 0, xend = PC1_arrow, yend = PC2_arrow),
               arrow = arrow(length = unit(0.3, "cm")),
               color = "black") +
  geom_text(data = loadings,
            aes(x = PC1_arrow * 1.1, y = PC2_arrow * 1.1, label = trait),
            color = "black", size = 3) +
  theme_bw() +
  labs(x = "PC1", y = "PC2"))
ggsave(here("Figures", "PCA", "pca loadings.medium.png"), plot = f2, width = 10, height =8)
```

```{r}
pca_scores <- as.data.frame(pca_result$x)
pca_scores$common_name <- traits$common_name
pca_scores$positive <- traits$positive
pca_scores$negative <- traits$negative
pca_scores$null <- traits$null

# 3️⃣ ggplot biplot colored by positive
(f3 <- ggplot(pca_scores, aes(x = PC1, y = PC2, color = factor(positive))) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_manual(values = c("0" = "grey", "1" = "#005AB5")) +
  theme_bw() +
  labs(color = "Positive Response", x = "PC1", y = "PC2",
       title = "PCA of Bird Traits Colored by Positive Fire Response"))
(f4 <- ggplot(pca_scores, aes(x = PC1, y = PC2, color = factor(negative))) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_manual(values = c("0" = "grey", "1" = "#DC3220")) +
  theme_bw() +
  labs(color = "Negative Response", x = "PC1", y = "PC2",
       title = "PCA of Bird Traits Colored by Negative Fire Response"))

(pos.neg <- f3/f4)
ggsave(here("Figures", "PCA", "pca by response.png"), plot = pos.neg, width = 8, height =10)


##
(f5 <- ggplot(pca_scores, aes(x = PC1, y = PC2, color = factor(positive))) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_manual(values = c("0" = "grey", "1" = "#005AB5")) +
  theme_bw() +
  geom_text(aes(label = common_name), size = 3) +
  labs(color = "Positive Response", x = "PC1", y = "PC2",
       title = "PCA of Bird Traits Colored by Positive Fire Response"))
(f6 <- ggplot(pca_scores, aes(x = PC1, y = PC2, color = factor(negative))) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_manual(values = c("0" = "grey", "1" = "#DC3220")) +
    geom_text(aes(label = common_name), size = 3) +
  theme_bw() +
  labs(color = "Negative Response", x = "PC1", y = "PC2",
       title = "PCA of Bird Traits Colored by Negative Fire Response"))

(pos.neg.spp.names <- f5/f6)
ggsave(here("Figures", "PCA", "pca by response spp names.medium.png"), plot = pos.neg.spp.names, width = 20, height =20, dpi = 300)

```

```{r}
# Extract loadings
loadings <- as.data.frame(pca_result$rotation)

# Add trait names as a column
loadings$trait <- rownames(loadings)

loadings_long <- loadings %>%
  pivot_longer(cols = starts_with("PC"), 
               names_to = "PC", 
               values_to = "loading") %>%
  filter(PC %in% paste0("PC", 1:2))   # keep only PCs 1-5

(loadings.bar <- ggplot(loadings_long, aes(x = trait, y = loading, fill = PC)) +
  geom_col(position = "dodge") +   # side-by-side bars per PC
  coord_flip() +                   # horizontal bars
  labs(x = "Trait", y = "Loading") +
  theme_bw(base_size = 14) +
  scale_y_continuous(limits = c(-0.5, 0.5), breaks = seq(-0.5, 0.5, 0.25)) +  # set range and breaks
  scale_fill_manual(values = c("#a1d99b", "#688c64")) +
  facet_wrap(~PC)+
    theme(legend.position = "none"))

ggsave(here("Figures", "PCA", "loadings bar plot.medium.png"), plot = loadings.bar, width = 6, height =5, dpi = 300)

loadings %>%  arrange(desc(abs(PC1)))


```
# Highest absolute loadings for PC! Positive - associated with negative responses, negative, associated with positive responses
habitat_density	-0.31424517		(lower number for habitat density indicates a density habitat, so dense? habitats contribute to postitive/null responses)
lifespan	-0.29169478	(lower lifespans are more positive/null responses)
live_tree_nesting_use	0.27513849	(higher live tree nesting associated with negative responses)
average_mass	-0.27241232	(lower mass are more positive/null responses)
ground_nester	-0.25813459	(ground nesters are associated with postitve responses)
```{r}
bc <- traits2 %>%  
  select(common_name, where(is.numeric), -elevational_range,  -kipps_distance, -repro_age, -clutches_per_year, -lifespan, -eco_spec_index, -ends_with("_id"), -fstimb) %>% 
  # mutate(null = NA,
  #        positive = NA,
  #        negative = NA) %>% 
  drop_na()  


#plotting all spp on this pca:
bc_scaled <- scale(bc[,-1],
                    center = (pca_result$center),
                    scale = (pca_result$scale))

bc_scores <- as.data.frame(as.matrix(bc_scaled) %*% pca_result$rotation)

```

```{r}

bc_scores$common_name <- bc$common_name
centers <- data.frame(
  group = c("positive", "mixed", "negative", "neutral"),
  PC1 = c(
    mean(pca_scores$PC1[pca_scores$positive == 1 & pca_scores$negative != 1], na.rm = TRUE),
    mean(pca_scores$PC1[pca_scores$positive == 1 & pca_scores$negative == 1], na.rm = TRUE),
    mean(pca_scores$PC1[pca_scores$negative == 1 & pca_scores$positive != 1], na.rm = TRUE),
    mean(pca_scores$PC1[pca_scores$null == 1 & pca_scores$positive == 0 & pca_scores$negative == 0], na.rm = TRUE)),
  PC2 = c(
    mean(pca_scores$PC2[pca_scores$positive == 1 & pca_scores$negative != 1], na.rm = TRUE),
    mean(pca_scores$PC2[pca_scores$positive == 1 & pca_scores$negative == 1], na.rm = TRUE),
    mean(pca_scores$PC2[pca_scores$negative == 1 & pca_scores$positive != 1], na.rm = TRUE),
    mean(pca_scores$PC2[pca_scores$null == 1 & pca_scores$positive == 0 & pca_scores$negative == 0], na.rm = TRUE)
  )
)

# Basic PCA scatterplot
(all <- ggplot(bc_scores, aes(x = PC1, y = PC2)) +  # V1 and V2 are PC1 and PC2
  geom_point(size = 3, alpha = 0.7) +
  labs(x = "PC1", y = "PC2", title = "All BC Birds Plotted on PCA Axes") +
  theme_bw())
ggsave(here("Figures", "PCA", "all spp.medium.png"), plot = all, width = 15, height =15, dpi = 300)


# Original PCA points
orig_df <- pca_scores
orig_df$type <- "PCA"


# New projected points
new_df <- bc_scores
new_df$common_name <- bc$common_name[complete.cases(bc)]

# Combine
plot_df <- left_join(new_df,orig_df)

# Overlay plot
(p1 <- ggplot(orig_df, aes(x = PC1, y = PC2)) +
    geom_point(data = new_df, aes( x = PC1, y = PC2), shape = 1, color = "grey", alpha = 0.85, size = 2)+
  geom_point(aes(color = factor(positive)), size = 2.5, alpha = 0.8) +
  scale_color_manual(values = c("0" = "grey50", "1" = "#005AB5")) +
  theme_bw() +
  labs(x = "PC1", y = "PC2", color = "Positive Response", shape = "Point Type"))

(n1 <- ggplot(orig_df, aes(x = PC1, y = PC2)) +
    geom_point(data = new_df, aes( x = PC1, y = PC2), shape = 1, color = "grey", alpha = 0.85, size = 2)+
  geom_point(aes(color = factor(negative)), size = 2.5, alpha = 0.8) +
  scale_color_manual(values = c("0" = "grey50", "1" = "#D55E00")) +
  theme_bw() +
  labs(x = "PC1", y = "PC2", color = "Negative Response", shape = "Point Type"))

orig_df <- orig_df %>% 
  mutate(
    medium = rowSums(data.frame(positive, negative), na.rm = TRUE),
    medium = ifelse(medium == 2|medium == 3, 1, 0))

(m1 <- ggplot(orig_df, aes(x = PC1, y = PC2)) +
    geom_point(data = new_df, aes( x = PC1, y = PC2), shape = 1, color = "grey", alpha = 0.85, size = 2)+
  geom_point(aes(color = factor(medium)), size = 2.5, alpha = 0.8) +
  scale_color_manual(values = c("0" = "grey50", "1" = "darkmagenta")) +
  theme_bw() +
  labs(x = "PC1", y = "PC2", color = "medium Response", shape = "Point Type"))

(pcas <- p1/n1/m1 + plot_annotation(tag_levels = 'A')
)
ggsave(here("Figures", "PCA", "PCA_all response patchwork.medium.png"), plot = pcas, width = 8, height =12, dpi = 300)


(with.mix <- ggplot(pca_scores) +
  geom_point(data = new_df, aes( x = PC1, y = PC2), shape = 1, color = "grey", alpha = 0.85, size = 2)+
  geom_point(aes(PC1, PC2), color = "grey50") +
  geom_point(data = subset(pca_scores, positive == 1),
             aes(PC1, PC2), color = "#0072B2", size = 3, alpha = 0.8) +
  geom_point(data = subset(pca_scores, negative == 1),
             aes(PC1, PC2), color = "#D55E00", size = 3, alpha = 0.8) +
  geom_point(data = subset(pca_scores, negative == 1& positive == 1),
             aes(PC1, PC2), color = "darkmagenta", size = 3, alpha = 0.8)+
  geom_point(data = centers, aes(PC1, PC2), size = 5, color = "white", shape = 16) +
  geom_point(data = centers, aes(PC1, PC2, color = group), size = 7, shape = 8) +
  scale_color_manual(values = c("darkmagenta", "#D55E00","grey50","#0072B2")) +
  theme_bw() +
  theme(legend.position="bottom")+
  labs(color = "Fire Response",
       title = ""))
ggsave(here("Figures", "PCA", "PCA_centers.medium.png"), plot = with.mix, width = 8, height =8, dpi = 300)


# t.test(
#   pca_scores$PC1[pca_scores$positive == 1&pca_scores$negative != 1],
#   pca_scores$PC1[pca_scores$negative == 1&pca_scores$positive != 1]
# )
# 
# # PC2
# t.test(
#   pca_scores$PC2[pca_scores$positive == 1&pca_scores$negative != 1],
#   pca_scores$PC2[pca_scores$negative == 1&pca_scores$positive != 1]
# )
#p < 0.05 → The positive and negative 1’s occupy significantly different mean positions along that PC.
#p ≥ 0.05 → No significant difference.
pca_scores <- pca_scores %>%
  mutate(fire_response = case_when(
    positive == 1 & negative == 1 ~ "mixed",
    positive == 1 & negative == 0 ~ "positive",
    negative == 1 & positive == 0 ~ "negative",
    TRUE ~ "neutral"
  ))
anova_pc1 <- aov(PC1 ~ fire_response, data = pca_scores)
summary(anova_pc1)

# Post-hoc (Tukey) test for pairwise comparisons
TukeyHSD(anova_pc1)

anova_pc2 <- aov(PC2~ fire_response, data = pca_scores)
summary(anova_pc2)

# Post-hoc (Tukey) test for pairwise comparisons
TukeyHSD(anova_pc2)


arrow_multiplier <- 5  # adjust to make arrows visible
loadings <- loadings %>%
  mutate(PC1_arrow = PC1 * arrow_multiplier,
         PC2_arrow = PC2 * arrow_multiplier)
(f2 <- ggplot() +
  geom_segment(data = loadings,
               aes(x = 0, y = 0, xend = PC1_arrow, yend = PC2_arrow),
               arrow = arrow(length = unit(0.3, "cm")),
               color = "black") +
   geom_text_repel(data = loadings,
                  aes(x = PC1_arrow, y = PC2_arrow, label = trait),
                  color = "black",
                  size = 4,
                  segment.color = "grey50",    # line from label to point
                  segment.size = 0.5,
                  segment.alpha = 0.8,
                  nudge_x = 0,                 # optional small nudge
                  nudge_y = 0,
                  box.padding = 0.5,
                  point.padding = 0.5,
                  max.overlaps = 20) +
  geom_point(data = centers, aes(PC1, PC2, color = group), size = 6, shape = 8) +
  theme_bw() +
  theme(legend.position = "none")+
  scale_color_manual(values = c("darkmagenta", "#D55E00","grey50","#0072B2")) +
  labs(x = paste0("PC1 (", round(summary(pca_result)$importance[2,1]*100, 1), "%)"),
    y = paste0("PC2 (", round(summary(pca_result)$importance[2,2]*100, 1), "%)")))
(medium <- with.mix+f2+  
  plot_layout(guides = "collect") +
  plot_annotation(theme = theme(legend.position = "bottom"), tag_levels = "A"))
ggsave(here("Figures", "PCA", "PCA_end fig medium.png"), plot = medium, width = 18, height =8, dpi = 300)

```

anova_pc1 <- aov(PC1 ~ fire_response, data = pca_scores)
> summary(anova_pc1)
               Df Sum Sq Mean Sq F value   Pr(>F)    
fire_response   3   95.2   31.75   6.353 0.000488 ***
Residuals     122  609.7    5.00                     
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
> 
> # Post-hoc (Tukey) test for pairwise comparisons
> TukeyHSD(anova_pc1)
  Tukey multiple comparisons of means
    95% family-wise confidence level

Fit: aov(formula = PC1 ~ fire_response, data = pca_scores)

$fire_response
                        diff        lwr        upr     p adj
negative-mixed     0.1054934 -1.8952531  2.1062399 0.9990705
neutral-mixed     -2.2028776 -3.9193403 -0.4864150 0.0059884
positive-mixed    -1.2816869 -2.9502444  0.3868705 0.1933438
neutral-negative  -2.3083710 -3.9548139 -0.6619281 0.0021519
positive-negative -1.3871803 -2.9836183  0.2092577 0.1124493
positive-neutral   0.9211907 -0.3002313  2.1426127 0.2071948

> 
> anova_pc2 <- aov(PC2~ fire_response, data = pca_scores)
> summary(anova_pc2)
               Df Sum Sq Mean Sq F value  Pr(>F)   
fire_response   3   49.6  16.543   5.653 0.00116 **
Residuals     122  357.0   2.926                   
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
> 
> # Post-hoc (Tukey) test for pairwise comparisons
> TukeyHSD(anova_pc2)
  Tukey multiple comparisons of means
    95% family-wise confidence level

Fit: aov(formula = PC2 ~ fire_response, data = pca_scores)

$fire_response
                         diff        lwr        upr     p adj
negative-mixed    -1.39758801 -2.9285988  0.1334228 0.0868184
neutral-mixed     -1.93872412 -3.2521953 -0.6252529 0.0010959
positive-mixed    -1.85339025 -3.1302035 -0.5765770 0.0013738
neutral-negative  -0.54113612 -1.8010268  0.7187546 0.6786873
positive-negative -0.45580224 -1.6774282  0.7658237 0.7656799
positive-neutral   0.08533388 -0.8493224  1.0199902 0.9952436