---
title: "species_groups"
author: "Jenna Loesberg"
date: "`r Sys.Date()`"
output:
  word_document: default
  html_document: default
---
```{r, include=FALSE}
library(here)
library(tidyverse)
```


```{r, echo = FALSE}
traits <- read.csv(here("Data", "cleaned", "BC_bird_traits.csv"))
```

Figure out which BC species are listed under SARA: 
list accessed from https://laws.justice.gc.ca/eng/acts/s-15.3/page-10.html
Below lists are all of the Canadian SARA spp, which are matched with the BC bird list
```{r SARA spp}
endangered <- c(
"Northern Bobwhite",
"Yellow-breasted Chat" ,#auricollis subspecies, Southern Mountain population and virens subspecies)
"Whooping Crane",
"Eskimo Curlew",
"Acadian Flycatcher",
"Horned Grebe", #(Magdalen Islands population)
"Ivory Gull",
"Red Knot",#subspecies (Southeastern U.S. / Gulf of Mexico / Caribbean wintering population and Tierra del Fuego and Patagonia wintering population)
"Streaked Horned Lark",
"Chestnut-collared Longspur",
"American Barn Owl",# (Eastern population)",  --> this should be the same as Tyto furcata but check (western is threatened, just listed as barn owl for SARA)
"Burrowing Owl",
"Spotted Owl",# (caurina subspecies)",
"Mountain Plover",
"Piping Plover",#(circumcinctus subspecies) and melodus subspecies
"King Rail",
"Greater Sage-Grouse", #(urophasianus subspecies)
"Williamson's Sapsucker",
"Pink-footed Shearwater",
"Eastern Loggerhead Shrike", # subspecies
"Coastal Vesper Sparrow",
"Henslow's Sparrow",
"Black Swift",
"Roseate Tern",
"Sage Thrasher",
"Cerulean Warbler",
"Kirtland's Warbler",
"Prothonotary Warbler",
"Red-headed Woodpecker",
"White-headed Woodpecker")
#these are for Canada-wide, which of these are not in the BC list of birds"
missing_names <- endangered[!(endangered %in% traits$common_name)] #do all these names appear in the BC list?
missing_names#these are the SARA spp not found in BC:

threatened <- c("Short-tailed Albatross",
"Least Bittern",
"Bobolink",
"Lark Bunting",
"Red Crossbill",# percna subspecies",
"American Goshawk",#laingi subspecies; SARA has "northern goshawk" listed but the spp name changed and I think this is the correct common name now
"Ross's Gull",
"Ferruginous Hawk",
"Red Knot",# roselaari subspecies",
"McCown's Longspur",
"Eastern Meadowlark",
"Marbled Murrelet",
"American Barn Owl",#(Western population) --> this should be the same as Tyto furcata
"Northern Saw-whet Owl", #brooksi subspecies
"Sprague's Pipit",
"Western Screech-Owl",#kennicottii subspecies and macfarlanei subspecies
"Loggerhead Shrike",#Prairie subspecies
"Bank Swallow",
"Barn Swallow",
"Chimney Swift",
"Bicknell's Thrush",
"Wood Thrush",
"Canada Warbler",
"Golden-winged Warbler",
"Louisiana Waterthrush",
"Eastern Whip-poor-will",
"Lewis's Woodpecker")
#these are for Canada-wide, which of these are not in the BC list of birds"
missing_names <- threatened[!(threatened %in% traits$common_name)] #do all these names appear in the BC list?
missing_names#these are the SARA spp not found in BC:

special.concern <- c("Black-footed Albatross",
"Cassin's Auklet",
"Rusty Blackbird",
"Long-billed Curlew",
"Harlequin Duck",# (Eastern population)",
"Peregrine Falcon", #pealei subspecies",
"Olive-sided Flycatcher",
"Barrow's Goldeneye",# (Eastern population)",
"Horned Grebe",# (Western population)",
"Western Grebe",
"Evening Grosbeak",
"Great Blue Heron",
"Red Knot",# rufa subspecies (Northeastern South America wintering population)",
"Ancient Murrelet",
"Common Nighthawk",
"Flammulated Owl",
"Short-eared Owl",
"Red-necked Phalarope",
"Band-tailed Pigeon",
"Yellow Rail",
"Buff-breasted Sandpiper",
"Baird's Sparrow",
"Harris's Sparrow",
"Grasshopper Sparrow",# pratensis subspecies",
"Savannah Sparrow", # princeps subspecies",
"Eastern Wood-pewee"
)
#these are for Canada-wide, which of these are not in the BC list of birds"
missing_names <- special.concern[!(special.concern %in% traits$common_name)] #do all these names appear in the BC list?
missing_names#these are the SARA spp not found in BC:

#Adding the species that match the BC list as a column in the above df
traits$SARA.endangered <- ifelse(traits$common_name %in% endangered, 1, NA)
traits$SARA.threatened <- ifelse(traits$common_name %in% threatened, 1, NA)
traits$SARA.special.concern <- ifelse(traits$common_name %in% special.concern, 1, NA)
```
# Now forming the other groups of interest:
### What do the numbers for migration mean?
Migration == 1] <- "sedentary"
Migration == 2] <- "partial"
$Migration == 3] <- "migratory"
```{r}
#goal here is to make groups for grouping the distributions
#groups should be: species at risk, land birds, forest birds, insectivores (aerial insectivores?), grassland species, grassland specialists, long distance migrants, semi migratory birds

groups <- traits %>% 
  mutate(long.distance.migrants = ifelse(migration == 3, 1, 0),
         short.distance.migrants = ifelse(migration == 2, 1, 0),
         residents = ifelse(migration == 1, 1, 0),
         aerial.insectivores = ifelse(family == "Hirundinidae"|family == "Apodidae"|family == "Caprimulgidae"|family == "Tyrannidae", 1, 0), #swallows, swifts, nightjars, and flycatcher families, informed by the State of Canada's Birds report
         invertivores = ifelse(trophic_niche == "Invertivore", 1, 0),
         grassland.species = ifelse(primary_habitat == "Grassland", 1, 0),
         forest.species = ifelse(primary_habitat == "Forest", 1, 0),
         SARA = ifelse(traits$SARA.endangered == 1 |traits$SARA.threatened == 1 |traits$SARA.special.concern == 1, 1,0)) %>% 
  select(order, family, sci_name, common_name, long.distance.migrants, short.distance.migrants, residents,aerial.insectivores, invertivores, grassland.species, forest.species, SARA, SARA.endangered, SARA.threatened, SARA.special.concern) #only keep these columns

```

# Get lists of each group here:
```{r}
long.distance.migrants <- groups %>% 
  filter(long.distance.migrants == 1) %>% 
  select(order, family, sci_name, common_name)
long.distance.migrants %>%
  select(2:4) %>%
  print()

short.distance.migrants <- groups %>% 
  filter(short.distance.migrants == 1) %>% 
  select(order, family, sci_name, common_name)
short.distance.migrants %>%
  select(2:4) %>%
  print()

residents <- groups %>% 
  filter(residents == 1) %>% 
  select(order, family, sci_name, common_name)
residents %>%
  select(2:4) %>%
  print()

residents <- groups %>% 
  filter(residents == 1) %>% 
  select(order, family, sci_name, common_name)
residents %>%
  select(2:4) %>%
  print()

aerial.insectivores <- groups %>% 
  filter(aerial.insectivores == 1) %>% 
  select(order, family, sci_name, common_name)
aerial.insectivores %>%
  select(2:4) %>%
  print()

invertivores <- groups %>% 
  filter(invertivores == 1) %>% 
  select(order, family, sci_name, common_name)
invertivores %>%
  select(2:4) %>%
  print()

grassland.species <- groups %>% 
  filter(grassland.species == 1) %>% 
  select(order, family, sci_name, common_name)
grassland.species %>%
  select(2:4) %>%
  print()

forest.species <- groups %>% 
  filter(forest.species == 1) %>% 
  select(order, family, sci_name, common_name)
forest.species %>%
  select(2:4) %>%
  print()

#This is for any spp listed under SARA
SARA <- groups %>% 
  filter(SARA == 1) %>% 
  select(order, family, sci_name, common_name)
SARA %>%
  select(2:4) %>%
  print()

#just endangered:
SARA.endangered <- groups %>% 
  filter(SARA.endangered == 1) %>% 
  select(order, family, sci_name, common_name)
SARA.endangered %>%
  select(2:4) %>%
  print()

#just threatened:
SARA.threatened <- groups %>% 
  filter(SARA.threatened == 1) %>% 
  select(order, family, sci_name, common_name)
SARA.threatened %>%
  select(2:4) %>%
  print()

#just special concern
SARA.special.concern <- groups %>% 
  filter(SARA.special.concern == 1) %>% 
  select(order, family, sci_name, common_name)
SARA.special.concern %>%
  select(2:4) %>%
  print()

```





