---
title: "species_groups"
author: "Jenna Loesberg"
date: "`r Sys.Date()`"
output:
  html_document: default
  word_document: default
---
```{r, include=FALSE}
library(here)
library(tidyverse)
```

can be viewed at: https://htmlpreview.github.io/?https://github.com/jloesberg/BCBirdsandFire/blob/main/Scripts/species.groups.html
```{r, echo = FALSE}
traits <- read.csv(here("Data", "cleaned", "BC_bird_traits.csv"))
```

Step 1. Figure out which BC species are listed under SARA (accessed from https://laws.justice.gc.ca/eng/acts/s-15.3/page-10.html)
- The code below lists all Canadian SARA spp, which I then match to the BC bird list
```{r SARA spp}
endangered <- c(
"Northern Bobwhite",
"Yellow-breasted Chat" ,#auricollis subspecies, Southern Mountain population and virens subspecies)
"Whooping Crane",
"Eskimo Curlew",
"Acadian Flycatcher",
"Horned Grebe", #(Magdalen Islands population)
"Ivory Gull",
"Red Knot",#subspecies (Southeastern U.S. / Gulf of Mexico / Caribbean wintering population and Tierra del Fuego and Patagonia wintering population)
"Streaked Horned Lark",
"Chestnut-collared Longspur",
"American Barn Owl",# (Eastern population)",  --> this should be the same as Tyto furcata but check (western is threatened, just listed as barn owl for SARA)
"Burrowing Owl",
"Spotted Owl",# (caurina subspecies)",
"Mountain Plover",
"Piping Plover",#(circumcinctus subspecies) and melodus subspecies
"King Rail",
"Greater Sage-Grouse", #(urophasianus subspecies)
"Williamson's Sapsucker",
"Pink-footed Shearwater",
"Eastern Loggerhead Shrike", # subspecies
"Coastal Vesper Sparrow",
"Henslow's Sparrow",
"Black Swift",
"Roseate Tern",
"Sage Thrasher",
"Cerulean Warbler",
"Kirtland's Warbler",
"Prothonotary Warbler",
"Red-headed Woodpecker",
"White-headed Woodpecker")
#Which of these Canada-wide SARA birds are not in the BC list of birds"
missing_names <- endangered[!(endangered %in% traits$common_name)]
missing_names #these are the SARA spp not found in BC:

threatened <- c("Short-tailed Albatross",
"Least Bittern",
"Bobolink",
"Lark Bunting",
"Red Crossbill",# percna subspecies",
"American Goshawk",#laingi subspecies; SARA has "northern goshawk" listed but the spp name changed and I think this is the correct common name now
"Ross's Gull",
"Ferruginous Hawk",
"Red Knot",# roselaari subspecies",
"McCown's Longspur",
"Eastern Meadowlark",
"Marbled Murrelet",
"American Barn Owl",#(Western population) --> this should be the same as Tyto furcata
"Northern Saw-whet Owl", #brooksi subspecies
"Sprague's Pipit",
"Western Screech-Owl",#kennicottii subspecies and macfarlanei subspecies
"Loggerhead Shrike",#Prairie subspecies
"Bank Swallow",
"Barn Swallow",
"Chimney Swift",
"Bicknell's Thrush",
"Wood Thrush",
"Canada Warbler",
"Golden-winged Warbler",
"Louisiana Waterthrush",
"Eastern Whip-poor-will",
"Lewis's Woodpecker")
#Which of these Canada-wide SARA birds are not in the BC list of birds"
missing_names <- threatened[!(threatened %in% traits$common_name)]
missing_names#these are the SARA spp not found in BC:

special.concern <- c("Black-footed Albatross",
"Cassin's Auklet",
"Rusty Blackbird",
"Long-billed Curlew",
"Harlequin Duck",# (Eastern population)",
"Peregrine Falcon", #pealei subspecies",
"Olive-sided Flycatcher",
"Barrow's Goldeneye",# (Eastern population)",
"Horned Grebe",# (Western population)",
"Western Grebe",
"Evening Grosbeak",
"Great Blue Heron",
"Red Knot",# rufa subspecies (Northeastern South America wintering population)",
"Ancient Murrelet",
"Common Nighthawk",
"Flammulated Owl",
"Short-eared Owl",
"Red-necked Phalarope",
"Band-tailed Pigeon",
"Yellow Rail",
"Buff-breasted Sandpiper",
"Baird's Sparrow",
"Harris's Sparrow",
"Grasshopper Sparrow",# pratensis subspecies",
"Savannah Sparrow", # princeps subspecies",
"Eastern Wood-pewee"
)
#Which of these Canada-wide SARA birds are not in the BC list of birds"
missing_names <- special.concern[!(special.concern %in% traits$common_name)]
missing_names#these are the SARA spp not found in BC:

#Adding the species that match the BC list as a column in the above df
traits$SARA.endangered <- ifelse(traits$common_name %in% endangered, 1, NA)
traits$SARA.threatened <- ifelse(traits$common_name %in% threatened, 1, NA)
traits$SARA.special.concern <- ifelse(traits$common_name %in% special.concern, 1, NA)
```
Step 2. Get BC listed species, accessed from https://a100.gov.bc.ca/pub/eswp/search.do with a search for all bird species in BC listed with a red or blue status
```{r BC listed spp}
blue <- c("Smith's Longspur",
"Rough-legged Hawk",
"American Golden-Plover",
"Gyrfalcon",
"Eared Grebe",
"Horned Lark",
"Black Swift",
"White-throated Swift",
"Baltimore Oriole",
"Winter Wren",
"Common Nighthawk",
"Black Tern",
"Killdeer",
"Rusty Blackbird",
"American Avocet",
"American Bittern",
"Lark Sparrow",
"Short-eared Owl",
"Surf Scoter",
"Red-necked Phalarope",
"American Barn Owl",
"Caspian Tern",
"Green Heron",
"Lewis's Woodpecker",
"Williamson's Sapsucker",
"Great Blue Heron",#, herodias subspecies",
"Canyon Wren",
"Western Screech-Owl",#, macfarlanei subspecies",
"Gray Flycatcher",
"Flammulated Owl",
"Brewer's Sparrow",#, breweri subspecies",
"Sharp-tailed Grouse",#, columbianus subspecies",
"Canada Warbler",
"Black-throated Green Warbler",
"Purple Martin",
"Double-crested Cormorant",
"Brant",
"Wandering Tattler",
"Connecticut Warbler",
"American Goshawk",#, atricapillus subspecies",
"Cape May Warbler",
"Black Scoter",
#"Great Blue Heron",#, fannini subspecies",
"Red Knot",
"Peregrine Falcon",#, pealei subspecies",
"Tufted Puffin",
"Ancient Murrelet",
"Band-tailed Pigeon",
"Marbled Murrelet",
"Northern Pygmy-Owl",#, swarthi subspecies",
"Western Screech-Owl",#, kennicottii subspecies",
"Northern Saw-whet Owl",#, brooksi subspecies",
"Steller's Jay",#, carlottae subspecies",
"Pine Grosbeak",#, carlottae subspecies",
"White-tailed Ptarmigan",#, saxatilis subspecies",
"Long-tailed Duck",
"Buller's Shearwater",
"Flesh-footed Shearwater",
"Pink-footed Shearwater",
"White-rumped Sandpiper",
"Stilt Sandpiper",
"Rock Sandpiper",
"Tundra Swan",
"Yellow-billed Loon",
"Western Gull",
"Marbled Godwit",
"Laysan Albatross",
"Black-footed Albatross",
"South Polar Skua",
"Common Tern",
"Lesser Yellowlegs")
#Which of these BC listed birds are not in our BC list of birds"
missing_names <- blue[!(blue %in% traits$common_name)]
missing_names #these are the SARA spp not found in BC:

red <- c(
"Greater Sage-Grouse",
"Prairie Falcon",
"Western Grebe",
"American White Pelican",
"California Gull",
"Forster's Tern",
"Bobolink",
"Short-billed Dowitcher",
"Upland Sandpiper",
"Peregrine Falcon",#,, anatum subspecies",
"Swainson's Hawk",
"Black-crowned Night-Heron",
"Sage Thrasher",
"Yellow-breasted Chat",
"Yellow-billed Cuckoo",
"Grasshopper Sparrow",
"Burrowing Owl",
"White-headed Woodpecker",
"Horned Lark",#, merrilli subspecies",
"Parasitic Jaeger",
"Hudsonian Godwit",
"Nelson's Sparrow",
"Bay-breasted Warbler",
"Yellow Rail",
"Vesper Sparrow",#, affinis subspecies",
"Brandt's Cormorant",
"American Goshawk",#, laingi subspecies",
"Horned Lark",#, strigata subspecies",
"Horned Puffin",
"Northern Fulmar",
"Cassin's Auklet",
"Common Murre",
"Thick-billed Murre",
"Spotted Owl",
"Clark's Grebe",
"Ruby-throated Hummingbird",
"Canada Goose",#, occidentalis subspecies",
"Baird's Sandpiper",
"Black-necked Stilt",
"Whimbrel",
"Short-tailed Albatross",
"Black-legged Kittiwake"
)
#Which of these BC listed birds are not in our BC list of birds"
missing_names <- red[!(red %in% traits$common_name)]
missing_names #these are the SARA spp not found in BC:

#Adding the species that match the BC list as a column in the above df
traits$blue.listed <- ifelse(traits$common_name %in% blue, 1, NA)
traits$red.listed <- ifelse(traits$common_name %in% red, 1, NA)
```



# Now forming the other groups of interest:
#### What do the numbers for migration mean?
Accessed from Avonet trait database

- Migration == 1] <- sedentary or residents
- Migration == 2] <- Partially migratory, i.e. minority of population migrates long distances, or most of population undergoes - short-distance migration, nomadic movements, distinct altitudinal migration, etc.
- Migration == 3] <- migratory, i.e. majority of population undertakes long-distance migration"

#### How were aerial insectivores grouped?
Informed by the State of Canada's Birds report, which assigns the swallow, swift, nightjar, and flycatcher families as aerial insectivores (accessed from https://naturecounts.ca/nc/socb-epoc/methods.jsp) 

#### Invertivores
Accessed from Avonet trait database

Invertivores are species obtaining at least 60% of food resources from invertebrates in terrestrial systems, including insects, worms, arachnids, etc.

#### Grassland and Forest species
Accessed from Birdbase trait database

Defined as main habitat type utilized by a species
    - Forest, secondary forest, swamp forest, tree-fall gaps, taiga, igapo, varzea
    -Grasslands, high altitude flats, puna, altiplano, tundra, tussock, pastures, steppe, paramo, moorland, plateaus, open vegetation

```{r}
#goal here is to make groups for grouping the distributions
#groups should be: species at risk, land birds, forest birds, insectivores (aerial insectivores?), grassland species, grassland specialists, long distance migrants, semi migratory birds

groups <- traits %>% 
  mutate(long.distance.migrants = ifelse(migration == 3, 1, 0),
         short.distance.migrants = ifelse(migration == 2, 1, 0),
         residents = ifelse(migration == 1, 1, 0),
         aerial.insectivores = ifelse(family == "Hirundinidae"|family == "Apodidae"|family == "Caprimulgidae"|family == "Tyrannidae", 1, 0), 
         invertivores = ifelse(trophic_niche == "Invertivore", 1, 0),
         grassland.species = ifelse(primary_habitat == "Grassland", 1, 0),
         forest.species = ifelse(primary_habitat == "Forest", 1, 0),
         SARA = ifelse(traits$SARA.endangered == 1 |traits$SARA.threatened == 1 |traits$SARA.special.concern == 1, 1,0)) %>% 
  select(order, family, sci_name, common_name, long.distance.migrants, short.distance.migrants, residents,aerial.insectivores, invertivores, grassland.species, forest.species, SARA, SARA.endangered, SARA.threatened, SARA.special.concern, blue.listed, red.listed) #only keep these columns

```

# Get lists of each group here:
```{r}
long.distance.migrants <- groups %>% 
  filter(long.distance.migrants == 1) %>% 
  select(order, family, sci_name, common_name)
long.distance.migrants %>%
  select(2:4) %>%
  print()

short.distance.migrants <- groups %>% 
  filter(short.distance.migrants == 1) %>% 
  select(order, family, sci_name, common_name)
short.distance.migrants %>%
  select(2:4) %>%
  print()

residents <- groups %>% 
  filter(residents == 1) %>% 
  select(order, family, sci_name, common_name)
residents %>%
  select(2:4) %>%
  print()

residents <- groups %>% 
  filter(residents == 1) %>% 
  select(order, family, sci_name, common_name)
residents %>%
  select(2:4) %>%
  print()

aerial.insectivores <- groups %>% 
  filter(aerial.insectivores == 1) %>% 
  select(order, family, sci_name, common_name)
aerial.insectivores %>%
  select(2:4) %>%
  print()

invertivores <- groups %>% 
  filter(invertivores == 1) %>% 
  select(order, family, sci_name, common_name)
invertivores %>%
  select(2:4) %>%
  print()

grassland.species <- groups %>% 
  filter(grassland.species == 1) %>% 
  select(order, family, sci_name, common_name)
grassland.species %>%
  select(2:4) %>%
  print()

forest.species <- groups %>% 
  filter(forest.species == 1) %>% 
  select(order, family, sci_name, common_name)
forest.species %>%
  select(2:4) %>%
  print()

#This is for any spp listed under SARA
SARA <- groups %>% 
  filter(SARA == 1) %>% 
  select(order, family, sci_name, common_name)
SARA %>%
  select(2:4) %>%
  print()

#just endangered:
SARA.endangered <- groups %>% 
  filter(SARA.endangered == 1) %>% 
  select(order, family, sci_name, common_name)
SARA.endangered %>%
  select(2:4) %>%
  print()

#just threatened:
SARA.threatened <- groups %>% 
  filter(SARA.threatened == 1) %>% 
  select(order, family, sci_name, common_name)
SARA.threatened %>%
  select(2:4) %>%
  print()

#just special concern
SARA.special.concern <- groups %>% 
  filter(SARA.special.concern == 1) %>% 
  select(order, family, sci_name, common_name)
SARA.special.concern %>%
  select(2:4) %>%
  print()

#BC blue listed
blue.listed <- groups %>% 
  filter(blue.listed == 1) %>% 
  select(order, family, sci_name, common_name)
blue.listed %>%
  select(2:4) %>%
  print()

#red listed
red.listed <- groups %>% 
  filter(red.listed == 1) %>% 
  select(order, family, sci_name, common_name)
red.listed %>%
  select(2:4) %>%
  print()

```





